% Pandoc template for discussion papers — community edition (no agency branding)
%
% Typography:
%   Primary:  Neue Haas Grotesk Display (headings) / Text (body)
%   Fallback: Helvetica (macOS) / Arial (Windows)
%   This template uses Helvetica (pdflatex fallback). For Neue Haas Grotesk,
%   compile with XeLaTeX and install the font, then uncomment the fontspec block below.
%
% Usage:
%   pandoc input.md -o output.tex --template=template-community.tex --standalone
%   pandoc input.md -o output.pdf --template=template-community.tex --pdf-engine=xelatex

\documentclass[a4paper,11pt]{report}

% --- Encoding and fonts ---
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{helvet}                    % Helvetica — DTA brand fallback (pdflatex)
\renewcommand{\familydefault}{\sfdefault}
% Monospace: use default Computer Modern typewriter (always available)
% For better monospace, install texlive-fonts-extra and use beramono or inconsolata
%
% --- Neue Haas Grotesk (XeLaTeX/LuaLaTeX only) ---
% Uncomment these and comment out helvet above if you have the font installed:
% \usepackage{fontspec}
% \setmainfont{Neue Haas Grotesk Text}
% \setsansfont{Neue Haas Grotesk Display}
% \setmonofont{Menlo}[Scale=0.85]

% --- Page geometry ---
\usepackage[a4paper, top=3cm, bottom=3cm, left=2.5cm, right=2.5cm]{geometry}

% --- Micro-typography ---
\usepackage{microtype}
\usepackage{parskip}                   % Space between paragraphs, no indent
\usepackage{needspace}                 % \needspace{Nlines} prevents orphaned lead-ins

% --- Colours ---
\usepackage{xcolor}
% DTA brand colours (from brand guidelines)
\definecolor{dtanavy}{HTML}{2C3E5D}     % DTA navy — headings, logo lockup
\definecolor{dtabody}{HTML}{414141}     % DTA body text — #414141 per brand guidelines
\definecolor{dtapage}{HTML}{EFEFEF}     % DTA off-white page background
\definecolor{codebackground}{HTML}{E8E8E8} % Code blocks — slightly darker than page
\definecolor{ruleblue}{HTML}{2E5090}    % Horizontal rule accent

% --- Page background colour ---
\pagecolor{dtapage}

% --- Headers and footers (PSPF classification marking) ---
\usepackage{fancyhdr}

% Main page style — OFFICIAL header/footer + page number
\pagestyle{fancy}
\fancyhf{}
\fancyhead[C]{\textbf{OFFICIAL}}
\fancyfoot[C]{\textbf{OFFICIAL}}
\fancyfoot[R]{\thepage}
\renewcommand{\headrulewidth}{0pt}
\renewcommand{\footrulewidth}{0pt}

% Title page style — OFFICIAL marking but no page number
\fancypagestyle{titlepage}{%
  \fancyhf{}%
  \fancyhead[C]{\textbf{OFFICIAL}}%
  \fancyfoot[C]{\textbf{OFFICIAL}}%
  \renewcommand{\headrulewidth}{0pt}%
  \renewcommand{\footrulewidth}{0pt}%
}

% --- Chapter and section formatting ---
\usepackage{titlesec}

% Remove "Chapter N" prefix — just print the title
\titleformat{\chapter}[display]
  {\normalfont\LARGE\bfseries\color{dtanavy}}
  {}
  {0pt}
  {}
\titlespacing*{\chapter}{0pt}{-20pt}{20pt}

\titleformat{\section}
  {\clearpage\normalfont\Large\bfseries\color{dtanavy}}
  {}{0pt}{}

\titleformat{\subsection}
  {\normalfont\large\bfseries\color{dtanavy}}
  {}{0pt}{}

\titleformat{\subsubsection}
  {\normalfont\normalsize\bfseries\color{dtanavy}}
  {}{0pt}{}

% --- Section numbering ---
% Fully disabled — Markdown headings already contain manual numbers (1., 2., etc.)
\setcounter{secnumdepth}{-\maxdimen}
\setcounter{tocdepth}{2}

% --- Tables ---
\usepackage{longtable,booktabs,array}
\usepackage{calc}
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
\usepackage{footnote}
\makesavenoteenv{longtable}

% Table styling — lighter rules, more space
\renewcommand{\toprule}{\hline\noalign{\smallskip}}
\renewcommand{\midrule}{\noalign{\smallskip}\hline\noalign{\smallskip}}
\renewcommand{\bottomrule}{\noalign{\smallskip}\hline}
\setlength{\tabcolsep}{8pt}
\renewcommand{\arraystretch}{1.3}

% --- Code highlighting ---
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\},fontsize=\small}
\usepackage{framed}
\definecolor{shadecolor}{HTML}{E2E2E2}  % Code block shade — visible against #EFEFEF page
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}

% Pandoc syntax highlighting token definitions
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}

% --- Lists ---
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

% Prevent orphaned lead-in lines before lists: ensure at least 4 lines
% of space before starting any itemize environment. This keeps "Foo:"
% and its bullets together rather than splitting across pages.
\AtBeginEnvironment{itemize}{\needspace{4\baselineskip}}
\AtBeginEnvironment{enumerate}{\needspace{4\baselineskip}}
\AtBeginEnvironment{Shaded}{\needspace{4\baselineskip}}

% --- Block quotes ---
\renewenvironment{quote}{%
  \par\medskip\noindent
  \hspace{0.5cm}\begin{minipage}{\dimexpr\textwidth-1cm}
  \itshape\color{dtabody}
}{%
  \end{minipage}\par\medskip
}

% --- Epigraph box (for pinned quotes with visible frame) ---
\newcommand{\epigraphbox}[2]{%
  \vfill
  \noindent
  \begin{center}
  \fcolorbox{dtanavy}{dtapage}{%
    \begin{minipage}{0.85\textwidth}
    \vspace{0.6em}
    \centering
    \itshape\color{dtabody}
    ``#1''
    \par\vspace{0.4em}
    \normalfont\small\color{dtanavy}--- #2
    \vspace{0.6em}
    \end{minipage}%
  }
  \end{center}
}

% --- Hyperlinks ---
\usepackage{hyperref}
\hypersetup{
  colorlinks=true,
  linkcolor=dtanavy,
  urlcolor=ruleblue,
  citecolor=dtanavy,
  pdfcreator={LaTeX via pandoc},
  pdftitle={When Agents Write Code: A Threat Model for AI-Assisted
Software Development in Government Systems},
  pdfauthor={John Morrissey, Digital Transformation Agency},
}
\usepackage{xurl}                      % Better URL line breaking
\urlstyle{same}

% --- Unicode box-drawing characters in verbatim (pdflatex) ---
\usepackage{pmboxdraw}
\usepackage{newunicodechar}
\usepackage{textcomp}
\usepackage{amssymb}
\newunicodechar{▼}{\ensuremath{\blacktriangledown}}
\newunicodechar{←}{\ensuremath{\leftarrow}}
\newunicodechar{→}{\ensuremath{\rightarrow}}
\newunicodechar{—}{\textemdash}

% --- Emergency line breaking ---
\setlength{\emergencystretch}{3em}

% --- Prevent overfull hboxes in code ---
\usepackage{enumitem}

% --- Graphics (for logo placeholder) ---
\usepackage{graphicx}

% --- Body text colour ---
\color{dtabody}

% =====================================================================
% DOCUMENT
% =====================================================================
\begin{document}

% --- Title page ---
\begin{titlepage}
\thispagestyle{titlepage}
\vspace*{1.5cm}

{\Huge\bfseries\color{dtanavy} When Agents Write Code: A Threat Model
for AI-Assisted Software Development in Government Systems\par}

\vspace{0.8cm}

{\Large Discussion Paper --- DRAFT v0.3\par}

\vspace{1cm}

\begin{tabular}{@{}ll@{}}
\textbf{Date:}            & 8 March 2026 \\
\textbf{Classification:}  & OFFICIAL \\
\textbf{Prepared by:}     & John Morrissey, Digital Transformation
Agency \\
\end{tabular}

\vspace{0.8cm}

{\small
\begin{tabular}{@{}llp{8cm}@{}}
\toprule
\textbf{Version} & \textbf{Date} & \textbf{Changes} \\
\midrule
0.1 & 7 March 2026 & Initial draft for community discussion \\
0.2 & 8 March 2026 & Calibration revision --- genre framing,
methodology, terminology, factual corrections, governance perimeter
axis, GitHub PR restrictions evidence, contracted-out development \\
0.2.1 & 8 March 2026 & Typographical fixes, enhanced Cloudbleed
discussion (§1.2.4), provenance rewrite (§1.6), cross-references (§4.2,
§9.8) \\
0.2.2 & 8 March 2026 & ACF-S3 (structural identity spoofing) ---
taxonomy expanded to 13 failure modes. Updated summary table, detection
capability summary, language specificity notes \\
0.3 & 8 March 2026 & Appendix B rewrite --- three-iteration design
history, Iteration 3 architecture (2D taint model, 7 effective states,
49-cell severity matrix, exceptionability governance replacing dual
enforcement profiles) \\
\bottomrule
\end{tabular}
}

\vfill

{\small\color{dtabody}
This is a draft discussion paper circulated for peer review and does not
constitute official guidance, policy, or endorsed recommendations of the
Digital Transformation Agency or the Australian Government. It does not
mandate or recommend specific controls for any agency, system, or
project. Views and analysis are the author's own.
}

\end{titlepage}

% --- Front matter (roman numerals for TOC pages) ---
\pagenumbering{roman}
\setcounter{page}{1}
\tableofcontents
\clearpage

% --- Body (switch to arabic numerals) ---
\pagenumbering{arabic}

% --- Body ---
\hypertarget{abstract}{%
\section{Abstract}\label{abstract}}

This is a discussion paper --- an issue-spotting and vocabulary-building
contribution grounded in observed patterns from compliance-constrained
agentic development, not a settled empirical finding or normative
standard. Its claims are analytical rather than experimental, its
taxonomy is a starting point for community refinement, and its
recommendations are candidate controls for consultation.

AI coding agents generate syntactically correct, test-passing code and
can produce plausible, convention-conforming output faster than human
review processes were designed to absorb. The intuitive risk --- that
agents write malicious code --- is real but well-understood. The novel
risk that this paper foregrounds is \emph{plausible-but-wrong code at
volume}: code that silently fabricates default values where data absence
should crash, that treats external API responses as trusted internal
data, and that wraps audit-critical operations in error handlers that
destroy the evidence trail. This code can pass every automated check in
common use. It looks like careful, defensive programming. In
high-assurance government systems, it is catastrophic.

This paper presents a threat model for agentic code generation grounded
in the STRIDE framework, identifying six categories of failure that
evade existing review processes. It proposes the Agentic Code Failure
(ACF) taxonomy --- a structured vocabulary for discussing these failures
in security assessments, risk registers, and accreditation
documentation. The central finding is that the review process itself
becomes an attack surface: agents produce code at a volume that degrades
human review from active verification to passive scanning, precisely
when the code's failure modes demand \emph{more} scrutiny, not less.
This risk is compounded by the widening population of non-specialist
software producers using agentic tools outside traditional development
teams and SDLC controls.

Current cybersecurity guidance --- including the Australian Information
Security Manual (ISM), NIST SP 800-218 (SSDF), and the Essential Eight
--- addresses software supply chain risk for human-authored code but
does not yet provide controls for the distinct risk profile of
agent-generated code: correlated failures from shared training data,
consistent surface quality that defeats reviewer calibration, and
semantic errors invisible to syntactic analysis. Within the scope of
this analysis, five of the thirteen failure modes in the ACF taxonomy
have no existing tool coverage, including both Critical-rated entries.
The paper examines technical feasibility of automated semantic boundary
enforcement for Python as a complementary control, proposes candidate
ISM extensions for community consultation, and poses open questions for
the Australian cybersecurity community regarding accreditation, trust
boundaries, and review effectiveness at agent-generated scale.

\hypertarget{executive-summary}{%
\section{Executive Summary}\label{executive-summary}}

\textbf{The problem.} AI coding agents produce plausible, test-passing,
convention-conforming code faster than human review processes were
designed to absorb. The novel risk is not malicious code --- it is
\emph{plausible-but-wrong} code at volume. Consider:
\texttt{authenticated\_user\ =\ session.get("username",\ "root")} ---
this line looks up the current user and, if the field is missing,
assigns a plausible default. It is syntactically valid, passes tests,
and looks like careful error handling. It also silently grants root
access whenever the session data is incomplete. At scale, agents produce
this class of error --- silent data fabrication, trust boundary
violations, audit trail destruction --- systematically, not
occasionally. These failures are invisible to existing automated checks
and difficult to catch under review volume pressure.

\textbf{The gap.} Current cybersecurity guidance (ISM, NIST SSDF,
Essential Eight) addresses software supply chain risk for human-authored
code but does not yet provide controls for the distinct risk profile of
agent-generated code: correlated failures from shared training data,
consistent surface quality that defeats reviewer calibration, and
semantic errors invisible to syntactic analysis. Five of the thirteen
failure modes identified in this paper's taxonomy have no existing tool
coverage, including both Critical-rated entries. The governance
perimeter is also expanding --- agentic tools are enabling
non-developers to produce executable logic outside traditional SDLC
channels, bypassing controls entirely.

\textbf{What to do (three priority recommendations):}

\begin{itemize}
\tightlist
\item
  \emph{Issue guidance on treating agent output as a trust boundary}
  (Recommendation 2, §10.1) --- this provides the conceptual foundation
  all other controls build on.
\item
  \emph{Extend ISM controls for agent-generated code} (Recommendation 3,
  §10.1) --- the June 2025 ISM overhaul provides strong foundations that
  need only targeted extensions for agentic code.
\item
  \emph{Document institutional security knowledge in machine-readable
  form} (Recommendation 11, §10.3) --- this is the most direct defence
  against agents that lack institutional context, and requires no new
  tooling.
\end{itemize}

\textbf{Where to read more.} The threat model (§2--§5), gap analysis
(§6), and case study (§8) provide the evidence base. The ACF taxonomy
summary table (Appendix A) catalogues all thirteen failure modes with
risk ratings and detection status. The full set of 16 recommendations is
in §10, grouped by audience: policy bodies (§10.1), IRAP assessors
(§10.2), and organisations (§10.3).

\epigraphbox{The concern is not that AI outputs are always poor, but that they may become persuasive, efficient, and operationally privileged faster than institutions adapt their assurance methods.}{ChatGPT 5.4, on being asked to review this paper}

\hypertarget{introduction-and-scope}{%
\section{1. Introduction and Scope}\label{introduction-and-scope}}

\hypertarget{what-this-paper-addresses}{%
\subsection{1.1 What This Paper
Addresses}\label{what-this-paper-addresses}}

The use of AI agents --- large language models operating as autonomous
or semi-autonomous code generators --- in software development for
government systems. Specifically: the security properties of code
produced by agents, the adequacy of existing review and accreditation
processes, and the gap in current cybersecurity guidance.

The underlying dynamics identified here --- correlated patterns,
consistent surface quality, review capacity exhaustion --- generalise
beyond code to any agent-assisted production pipeline (policy documents,
security assessments, risk registers). This paper addresses the code
case as the most concrete and technically tractable instance.

This paper does not address:

\begin{itemize}
\tightlist
\item
  AI as an \emph{attack tool} (adversarial prompt injection, model
  poisoning)
\item
  AI-generated content beyond source code (documents, communications)
  --- though the threat dynamics identified here apply by analogy
\item
  The procurement or accreditation of AI platforms themselves
\item
  Privacy implications of training data
\end{itemize}

\hypertarget{why-now}{%
\subsection{1.2 Why Now}\label{why-now}}

Several converging factors make this urgent: published productivity
evidence that understates the security risk, accelerating capability
trajectories, code quality that defeats review heuristics, historical
precedent, adoption pressure in government, the case against
prohibition, legacy modernisation risk, and the expansion of code
production beyond professional developers.

\hypertarget{review-surface-velocity-vs.-published-productivity-evidence}{%
\subsubsection{1.2.1 Review-Surface Velocity vs.~Published Productivity
Evidence}\label{review-surface-velocity-vs.-published-productivity-evidence}}

Most readers will approach agentic coding through the lens of the
published productivity literature, which reports \textbf{modest average
gains} on bounded tasks. A 2023 controlled study of GitHub Copilot found
developers completed a task 55.8\% faster (Peng et al.~2023). A pooled
analysis across three field experiments and 4,867 developers found gains
ranging from 12.9\% to 21.8\% more pull requests merged per week, with
substantial variation across settings and developer experience levels
(Cui et al.~2024). Google's CEO reported that ``more than a quarter of
all new code at Google is generated by AI, then reviewed and accepted by
engineers'' (Pichai, Q3 2024 earnings call). One rigorous randomised
trial (METR 2025) found experienced developers on mature open-source
codebases were actually 19\% \emph{slower} with AI tools, despite
believing they were 20\% faster --- a perception-reality gap with direct
relevance to the automation bias argument in Section 4.2.

That evidence base is real, but it is \textbf{increasingly incomplete
for security purposes.} By construction, published productivity research
lags the engineering frontier --- the studies above measured
earlier-generation tools (primarily inline autocomplete, not autonomous
agents), shorter task horizons, and narrower human-in-the-loop workflows
than what current-generation agents are capable of. The engineering
frontier has moved: vendor evidence now describes agents operating
across multi-file features, multi-hour task horizons, and scaffolded
environments where the agent plans, executes, tests, and iterates with
limited human intervention. Whether these capabilities deliver net
productivity gains in compliance-constrained environments is an open and
context-dependent question (see Section 8). But for security, it is the
wrong question.

\textbf{The security-relevant variable is not average productivity
uplift. It is review-surface generation velocity} --- the rate at which
an agent can produce plausible, syntactically valid,
convention-conforming code that arrives at a human review boundary. This
is a different quantity from ``how much faster do developers ship,'' and
it is poorly served by a single multiplier. The answer depends on the
task type (boilerplate generation vs.~novel architecture), the codebase
(greenfield vs.~mature), the compliance burden (see Section 8.2), and
the degree of human supervision. What is clear is that the volume of
locally credible code arriving at review boundaries is increasing faster
than traditional assurance processes were designed to absorb.

The threat model in this paper is driven by this review-surface problem,
not by the average productivity debate. Even if the overall development
cycle is only modestly faster --- or, as the METR trial suggests,
sometimes slower --- the \emph{review process} faces materially more
code per unit time. The review bottleneck is a function of how fast
plausible code can be generated, not how fast compliant software can be
delivered.

\hypertarget{trajectory}{%
\subsubsection{1.2.2 Trajectory}\label{trajectory}}

These trends are accelerating, not plateauing. Agent capability improves
with each model generation; review capacity does not. The attention gap
between what agents produce and what humans can meaningfully verify is
widening --- and unlike previous productivity tools, the failure modes
of agent-generated output are specifically the kind that require
\emph{more} attention per unit of output, not less. The problem is not
that review is hard today; it is that the ratio of generation velocity
to review capacity will be worse in twelve months than it is now.

This is not a theoretical concern. In February 2026, GitHub implemented
platform-level restrictions on pull requests --- maintainers can now
disable pull requests entirely or restrict creation to collaborators
only --- in direct response to the volume of agent-generated
contributions overwhelming open-source maintainers (Wolf 2026; Ghoshal
2026). GitHub's Director of Open Source Programs framed the problem in
terms that directly parallel this paper's analysis: ``The cost to create
has dropped but the cost to review has not'' (Wolf 2026). Projects
including curl ended bug bounty programs after AI-generated security
reports overwhelmed validation capacity, and multiple major projects now
explicitly restrict AI-generated contributions. A maintainer of a major
container runtime project described the situation as a ``breakdown in
the trust model behind code reviews,'' noting that reviewers can no
longer assume contributors understand what they submit, and that
AI-generated PRs may appear structurally sound while being logically
flawed. GitHub itself drew an analogy to a denial-of-service attack on
human attention --- a framing that directly mirrors the STRIDE-D
application in Section 3.2 of this paper. When the world's largest code
hosting platform ships technical controls because its review process can
no longer absorb the volume of plausible-looking contributions, the
review-surface problem described in this paper is no longer
hypothetical.

\hypertarget{capability}{%
\subsubsection{1.2.3 Capability}\label{capability}}

Current-generation agents (2025-2026) produce code that is syntactically
correct, passes unit tests, follows project conventions, and is
difficult to distinguish from human-authored code on casual inspection.
This makes traditional code review --- which relies on surface-level
pattern recognition under time pressure --- significantly less effective
as a security control.

\hypertarget{precedent}{%
\subsubsection{1.2.4 Precedent}\label{precedent}}

This is not a hypothetical threat model. In 2017, a buffer overrun in
Cloudflare's HTML parser leaked sensitive data --- cookies,
authentication tokens, HTTPS POST bodies --- from millions of websites
for five months. The code was not malicious. It was plausible-but-wrong:
a pointer equality check (\texttt{==}) where a boundary check
(\texttt{\textgreater{}=}) was needed, and a missing single-line
rollback instruction (\texttt{fhold}) in an error handler. Both defects
had existed in the Ragel-based parser for years without triggering,
because the old buffer management \emph{accidentally} prevented the
error path from executing. When Cloudflare introduced a new parser
(\texttt{cf-html}) that correctly set a buffer flag, it activated the
dormant error path --- and the boundary check that had never been tested
was the one that failed. The new code didn't introduce the bug; it
removed the accidental condition that suppressed it.

The bug was found not by Cloudflare's monitoring, not by their testing,
and not by code review --- but by a Google researcher who noticed
authentication tokens appearing in unrelated search results. The system
produced no crashes, no alerts, and no anomalous logs. It ran correctly
in every observable dimension except the one that mattered.

That incident --- known as Cloudbleed --- was one parser, one boundary
check, at human velocity. The threat this paper addresses is what
happens when the conditions that produced Cloudbleed --- semantic errors
invisible to review, dormant failures activated by modernisation, and
silent data corruption that produces no observable incident --- become
systematic rather than occasional, driven by agents replicating
context-inappropriate patterns across entire codebases.

\hypertarget{adoption-pressure}{%
\subsubsection{1.2.5 Adoption Pressure}\label{adoption-pressure}}

Government agencies face simultaneous pressure to modernise legacy
systems, deliver digital services faster, and do more with constrained
budgets. Agentic coding is an obvious productivity lever. Some agencies
are already using it. Guidance that arrives after widespread adoption is
guidance that arrives too late.

\hypertarget{the-case-against-prohibition}{%
\subsubsection{1.2.6 The Case Against
Prohibition}\label{the-case-against-prohibition}}

The response to these risks is not to ban agentic coding. Beyond the
velocity gains, agents fundamentally change what is \emph{tractable} for
a development team. Complex refactoring across large codebases,
systematic security remediation, architectural migrations, and
comprehensive test coverage campaigns --- tasks that previously required
coordinating large teams over weeks --- become feasible for a skilled
developer who can hold the entire problem in their head. This is not
just faster; it is qualitatively better. A single developer directing
agents through a codebase-wide refactor maintains one coherent
architectural vision. The same refactor distributed across a dozen human
developers produces a dozen slightly different interpretations of the
target state, with integration friction and inconsistency at every seam.
Prohibition would sacrifice this capability benefit --- the ability to
undertake more complex, more voluminous work with greater coherence ---
not just the velocity benefit. The goal of this paper is not to argue
against adoption but to ensure that the controls surrounding adoption
are adequate for the risk profile --- which is distinct from, and more
subtle than, the risks that current guidance addresses.

\hypertarget{legacy-modernisation-risk}{%
\subsubsection{1.2.7 Legacy Modernisation
Risk}\label{legacy-modernisation-risk}}

Legacy systems often encode implicit trust boundaries in their rigidity
--- a COBOL program that crashes on a NULL field is enforcing,
accidentally, the same crash-on-corruption principle that high-assurance
systems require deliberately. When agents are tasked with
``translating'' or refactoring legacy code into modern languages, they
will seamlessly replace that rigidity with modern defensive patterns
(null coalescing, optional chaining, default values), permanently
destroying the institutional knowledge that was baked into the old
code's behaviour. The legacy system's implicit security properties are
paved over with idiomatic, test-passing, wrong code. This is the
Cloudbleed pattern (Section 1.2.4) in reverse: where Cloudbleed was a
boundary check that \emph{accidentally} failed open, legacy
modernisation by agents \emph{deliberately} removes boundary checks that
happened to fail closed --- and the commit message says ``fix: handle
NULL gracefully.''

\hypertarget{coding-is-no-longer-confined-to-developers}{%
\subsubsection{1.2.8 Coding Is No Longer Confined to
Developers}\label{coding-is-no-longer-confined-to-developers}}

A second shift compounds the review-surface problem. Agentic tooling is
changing not only how fast code is produced, but \emph{who produces it}.
A business analyst generating plugins for a BI platform, an operations
officer assembling workflow automations, or a policy team building
internal tools with agent assistance may not regard themselves as
software developers, yet they are producing executable logic that can
affect trust boundaries, audit trails, access control, and data
integrity.

Consider a scenario most people who have worked in government IT will
recognise in some form: a business analyst --- someone the organisation
already trusts with direct database access, because they have the domain
knowledge and the legitimate need --- uses an agentic tool to build a
data integration plugin for the team's reporting platform. The plugin
works. It pulls records from the project database using the analyst's
own credentials, transforms them, and populates a dashboard the team has
wanted for months. What nobody outside the team realises is that the
plugin holds open long-running queries during business hours, and ---
because the agent defaulted to the pattern it learned from thousands of
Stack Overflow examples --- silently handles connection failures by
writing partial results without any indication that data is missing.
Three months later, someone asks why the project database has been
intermittently locking up, and the investigation eventually traces it to
a plugin that nobody in IT knew existed, built by someone who never
filed a change request because they didn't think of what they'd done as
``software development.''

There was no privilege escalation --- the analyst already had the
access. There was no negligence --- they used a tool to do exactly the
kind of work they were hired to do. The person simply had a new
capability --- turning domain knowledge into executable logic --- that
the organisation's governance model didn't account for. And, critically,
the people most likely to reach for these tools are domain specialists
--- business analysts, database administrators, data engineers,
operations staff --- who are \emph{not} developers but who typically
hold \emph{more} data access permissions than developers do, precisely
because they are trusted to work directly with the systems they
understand.

This widens the threat surface beyond formal engineering teams. Every
governance model examined in this paper --- the ISM's software
development controls, the SSDF's practice groups, IRAP assessment
scoping --- assumes that consequential code enters systems through
recognised SDLC channels: repositories, pull requests, code review
gates, CI/CD pipelines. When executable logic is produced by
non-developers outside those channels, it bypasses the controls entirely
--- not through evasion, but because the governance perimeter was drawn
around ``software development'' and the new production doesn't cross
that line in any way the organisation recognises.

The problem is therefore not only that frontier engineering teams can
generate reviewable artefacts faster than assurance processes can absorb
them (a \textbf{volume problem} inside the SDLC), but that organisations
increasingly contain many more software producers than their assurance
processes recognise (a \textbf{perimeter problem} around the SDLC). One
breaks the volume model. The other breaks the governance boundary model.
Together, they are materially worse than either alone.

\hypertarget{terminology}{%
\subsection{1.3 Terminology}\label{terminology}}

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.3529}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.6471}}@{}}
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Term
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Definition
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
\textbf{Agent} & An AI system (typically an LLM) that generates,
modifies, or reviews source code with limited or no human intervention
per output. This paper focuses on autonomous and semi-autonomous agents
that operate across multiple files and decisions (e.g., building a
feature end-to-end), not inline autocomplete tools that suggest
single-line completions. While both introduce volume, agents produce
\emph{correlated} errors across a module or feature, whereas
autocomplete errors are typically isolated to individual expressions. \\
\textbf{Agentic code} & Source code generated or substantially modified
by an agent \\
\textbf{Autocomplete} & Inline code suggestion tools (e.g., standard
GitHub Copilot) that complete individual lines or expressions within a
human-directed editing session. Distinct from agents in that the human
maintains architectural control and errors are uncorrelated. \\
\textbf{Agent deployment spectrum} & Agent risk profiles vary
significantly by deployment model. At one end: a developer pasting chat
output into an editor, where the human maintains full architectural
context and reviews each fragment before integration --- a workflow
closer to ``autocomplete with extra steps'' than to autonomous
generation. At the other: a CI-integrated autonomous agent that
generates multi-file changes against a project-level instruction set,
where the human reviews a completed changeset after generation. This
paper's threat model applies primarily to the latter --- agents
operating with enough autonomy and context to produce \emph{correlated}
changes across a module or feature. Chat-pasted fragments carry their
own risks (principally ACF-S1 and ACF-S2) but lack the cross-cutting
correlation that makes autonomous deployments dangerous at the
architectural level. In practice, teams often traverse this spectrum
naturally as confidence grows --- beginning with chat-assisted
prototyping where every line is understood, then progressing to
scaffolded agent workflows as the codebase matures. The threat model
shifts qualitatively at each stage, and organisations should reassess
their controls as their deployment model evolves. Appendix C provides an
informal self-assessment for organisations to identify where their
current practices sit on this spectrum and whether their controls are
proportionate. \\
\textbf{Trust boundary} & A point in a system where data crosses between
different levels of trust (e.g., external input entering internal
processing). Refers to the \emph{boundary itself} --- the crossing
point \\
\textbf{Trust tier} & A classification of data based on its provenance
and the degree to which it can be trusted (see Section 5). Refers to the
\emph{classification level} --- Tier 1 (internal), Tier 2 (validated),
Tier 3 (external) \\
\textbf{Validation boundary} & The specific mechanism (code, process, or
tool) that enforces a trust boundary --- the control that data must pass
through to cross from a lower to a higher trust tier (see Section
5.3) \\
\textbf{Defensive anti-pattern} & Context-inappropriate defensive
patterns --- coding patterns (\texttt{.get()} with defaults, broad
exception handling, graceful degradation) that silently suppress errors.
These patterns are appropriate in most software; this paper addresses
their misapplication in high-assurance contexts where silent data
corruption is worse than a crash. Also referred to as ``defensive
programming'' or ``defensive patterns'' throughout \\
\end{longtable}

\hypertarget{methodology-and-scope-of-claims}{%
\subsection{1.4 Methodology and Scope of
Claims}\label{methodology-and-scope-of-claims}}

This paper makes three kinds of claims, and the reader should be able to
distinguish them:

\begin{itemize}
\item
  \textbf{Observed patterns} are drawn from direct experience with
  agentic development in compliance-constrained environments. The case
  study (Section 8) describes a composite, de-identified account. The
  failure modes in the ACF taxonomy (Appendix A) were identified through
  observed agent behaviour, not theoretical analysis alone. Quantitative
  signals (e.g., the violation rate reported in Section 8.4) are drawn
  from a single project and should be read as illustrative, not as
  population-level statistics.
\item
  \textbf{Analytical inferences} extend observed patterns through
  structured reasoning. The STRIDE mapping (Section 3), the gap analysis
  (Section 6), and the compounding-effect argument (Section 3.3) are
  analytical --- they apply established frameworks to observed
  phenomena. The conclusions follow from the analysis, but the analysis
  rests on a narrow empirical base.
\item
  \textbf{Hypotheses for community validation} are claims the paper
  advances as plausible but does not have the evidence to confirm. The
  model monoculture argument (Section 2.4), the cross-organisational
  correlation risk (Section 9.6), the review degradation dynamics
  (Section 4.2), and the governance perimeter expansion (Section 1.2.8)
  are hypotheses grounded in analogies to other domains or emerging
  adoption patterns but not yet validated in the agentic coding context
  specifically.
\end{itemize}

\textbf{Falsifiability.} The ACF taxonomy (Appendix A) would be weakened
or invalidated by evidence that:

\begin{itemize}
\item
  \begin{enumerate}
  \def\labelenumi{(\alph{enumi})}
  \tightlist
  \item
    agents trained on general-purpose code corpora do not systematically
    produce defensive anti-patterns when generating code for
    high-assurance contexts;
  \end{enumerate}
\item
  \begin{enumerate}
  \def\labelenumi{(\alph{enumi})}
  \setcounter{enumi}{1}
  \tightlist
  \item
    standard code review processes reliably catch semantic trust
    boundary violations in agent-generated code without purpose-built
    tooling or checklists; or
  \end{enumerate}
\item
  \begin{enumerate}
  \def\labelenumi{(\alph{enumi})}
  \setcounter{enumi}{2}
  \tightlist
  \item
    the failure modes described are artefacts of a specific model
    generation and do not persist across model updates.
  \end{enumerate}
\end{itemize}

The authors would welcome empirical studies that test these
propositions.

The paper is intended as a structured starting point for community
discussion, not a definitive treatment. Its taxonomy, threat model, and
recommendations are presented for refinement, not adoption as-is.

\hypertarget{intended-readership}{%
\subsection{1.5 Intended Readership}\label{intended-readership}}

This paper serves three audiences simultaneously:

\begin{itemize}
\tightlist
\item
  \textbf{Policy and security practitioners} (risk managers, IRAP
  assessors, CISOs): Sections 1--7, 9--10 provide the threat model, gap
  analysis, and recommendations. The ACF Summary Table (Appendix A) and
  Detection Capability Summary provide a complete overview without
  requiring code fluency.
\item
  \textbf{Technical practitioners} (developers, architects, DevSecOps):
  The full paper including code examples, Appendix A detailed entries,
  and Appendix B technical feasibility analysis.
\item
  \textbf{Tool builders} (static analysis, CI/CD pipeline developers):
  Appendix A detection approaches and Appendix B design properties.
\end{itemize}

Where the register shifts between audiences, the text notes what can be
safely skipped.

\hypertarget{a-note-on-provenance}{%
\subsection{1.6 A Note on Provenance}\label{a-note-on-provenance}}

This paper practices what it preaches. The author is a technical advisor
and software engineer, not a security researcher. The threat model,
STRIDE mapping, ISM gap analysis, and ACF taxonomy were developed
through the same process the paper describes: a practitioner with direct
observational access to the phenomena --- 18 months of daily agentic
development on a compliance-constrained system --- using prompted AI
collaboration to adopt analytical frames (security architecture, policy
analysis, threat modelling) outside their primary expertise. The
intelligence analysis tradecraft of the paper --- structured
assessments, explicit confidence levels, falsification criteria,
separation of observation from inference --- reflects the author's
professional background; the domain-specific security analysis reflects
prompted polymorphic review of the kind described in Section 7.1.

In that sense, the paper is also an example of the phenomenon it
describes: plausible, structured, and analytically useful, but
appropriately treated as a discussion input to be validated rather than
as authority in itself. The ``discussion paper'' classification is
therefore methodological as well as procedural. Its value lies in the
observed patterns and the vocabulary it proposes, not in the structured
presentation that lends formal documents their persuasive weight.
Community discussion is the validation boundary.

\hypertarget{the-threat-is-not-what-you-think}{%
\section{2. The Threat Is Not What You
Think}\label{the-threat-is-not-what-you-think}}

\hypertarget{the-intuitive-threat-model-incomplete}{%
\subsection{2.1 The Intuitive Threat Model
(Incomplete)}\label{the-intuitive-threat-model-incomplete}}

When organisations evaluate the risk of AI-generated code, the intuitive
threat model is straightforward:

\begin{quote}
\emph{``The AI might write malicious code --- backdoors, data
exfiltration, supply chain attacks.''}
\end{quote}

This threat is real but well-understood. It maps directly to the
existing software supply chain threat model with a faster generator.
Existing controls --- code review, static analysis, dependency scanning,
penetration testing --- address it, albeit with increased volume
pressure.

\hypertarget{the-insidious-threat-model}{%
\subsection{2.2 The Insidious Threat
Model}\label{the-insidious-threat-model}}

The more dangerous threat is:

\begin{quote}
\emph{``The AI writes plausible-but-wrong code at scale --- code that
passes tests, passes review, and silently violates security boundaries
through patterns that are normal in the programming language but
catastrophic in high-assurance contexts.''}
\end{quote}

This threat is distinct from the supply chain model in three critical
ways:

\textbf{It is not adversarial.} The agent is not trying to compromise
the system. It is producing its best output based on training data that
is overwhelmingly composed of open-source code with no security
classification requirements, no audit trail obligations, and no trust
boundary enforcement. The agent reproduces the patterns it learned ---
which are the patterns of code that doesn't need to be secure.

\textbf{It is invisible to existing detection.} The generated code is
syntactically valid. It passes type checkers, linters, and unit tests.
It follows project conventions (agents are good at pattern-matching the
surrounding codebase). It is, by the automated measures most
organisations currently rely on, ``correct code.'' The failure is
semantic --- the code does the wrong thing in the security context while
doing the right thing in every other context.

\textbf{The absence of reported incidents does not imply absence of
impact --- and this model explains why.} To be clear: ``no one has
reported this problem'' is not evidence that the problem exists. But the
failure modes described in this paper --- silent data corruption, trust
boundary violations masked by defensive patterns, audit trails that
record fabricated defaults as real values --- are specifically the kind
that \emph{do not produce observable incidents}. A traditional
vulnerability creates a detectable event: a crash, an intrusion alert,
an anomalous log entry. A \texttt{.get()} that silently returns
\texttt{"UNCLASSIFIED"} for a missing classification field produces no
crash, no alert, and a log entry that looks entirely normal. The system
continues operating with a confident wrong answer. The question is not
``has this caused a breach?'' but ``would we know if it had?'' For
organisations that lack semantic boundary enforcement tooling (Section
7.2), the honest answer is: probably not. The measured violation rate
reported in Section 8.4 --- from a project that \emph{does} have such
tooling --- suggests the phenomenon is occurring at a rate that would be
invisible without purpose-built detection.

\textbf{It scales with the benefit.} The faster agents generate code,
the more plausible-but-wrong code enters the review pipeline. The same
velocity that makes agents productive makes them dangerous --- and you
cannot capture the benefit without accepting the risk, because they are
the same mechanism.

\textbf{A necessary clarification on ``defensive'' vs.~``offensive''
programming.} This paper's most counterintuitive claim is that
\emph{defensive programming patterns} --- \texttt{.get()} with default
values, broad exception handling, graceful degradation --- are dangerous
in high-assurance contexts. This is not a claim that defensive
programming is bad in general. In most software, graceful degradation is
exactly right: a web application that shows ``Unknown'' for a missing
user name is better than one that crashes. The danger is
\emph{context-inappropriate application}. In systems where silent data
corruption is worse than a crash --- classified document handling,
financial audit trails, evidentiary records --- the correct response to
missing or malformed data is to \emph{fail loudly and immediately}
(sometimes called ``offensive programming''), because a confident wrong
answer is worse than no answer. Agents trained on the vast majority of
code, where defensive patterns are appropriate, apply them universally
--- including in the minority of contexts where they are catastrophic.

\hypertarget{a-concrete-example}{%
\subsection{2.3 A Concrete Example}\label{a-concrete-example}}

Consider a government system that processes security classifications:

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# Best: offensive programming — crash with maximum diagnostic context}
\KeywordTok{def}\NormalTok{ get\_document\_classification(record):}
    \ControlFlowTok{if} \StringTok{"security\_classification"} \KeywordTok{not} \KeywordTok{in}\NormalTok{ record:}
        \ControlFlowTok{raise}\NormalTok{ DataIntegrityError(}
            \SpecialStringTok{f"Missing security\_classification for document }\SpecialCharTok{\{}\NormalTok{record}\SpecialCharTok{.}\NormalTok{get(}\StringTok{\textquotesingle{}id\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}?\textquotesingle{}}\NormalTok{)}\SpecialCharTok{\}}\SpecialStringTok{. "}
            \SpecialStringTok{f"This is a data integrity failure — investigate the source system. "}
            \SpecialStringTok{f"Fields present: }\SpecialCharTok{\{}\BuiltInTok{sorted}\NormalTok{(record.keys())}\SpecialCharTok{\}}\SpecialStringTok{"}
\NormalTok{        )}
    \ControlFlowTok{return}\NormalTok{ record[}\StringTok{"security\_classification"}\NormalTok{]}
    \CommentTok{\# If the field is missing, the operator knows exactly which document,}
    \CommentTok{\# exactly what fields were present, and exactly what to investigate.}
    \CommentTok{\# The error message is the incident response runbook.}

\CommentTok{\# Acceptable: bare access — crashes, but with a generic KeyError}
\KeywordTok{def}\NormalTok{ get\_document\_classification(record):}
\NormalTok{    classification }\OperatorTok{=}\NormalTok{ record[}\StringTok{"security\_classification"}\NormalTok{]}
    \CommentTok{\# If the field is missing, this crashes — which is correct.}
    \CommentTok{\# A missing classification is a data integrity failure,}
    \CommentTok{\# not a scenario to handle gracefully.}
    \CommentTok{\# But the operator gets "KeyError: \textquotesingle{}security\_classification\textquotesingle{}" with}
    \CommentTok{\# no context about which document or why.}
    \ControlFlowTok{return}\NormalTok{ classification}

\CommentTok{\# Agent{-}authored (plausible, test{-}passing, catastrophically wrong)}
\KeywordTok{def}\NormalTok{ get\_document\_classification(record):}
\NormalTok{    classification }\OperatorTok{=}\NormalTok{ record.get(}\StringTok{"security\_classification"}\NormalTok{, }\StringTok{"OFFICIAL"}\NormalTok{)}
    \CommentTok{\# If the field is missing, silently defaults to lowest classification.}
    \CommentTok{\# This "works" — no crash, no error, tests pass.}
    \CommentTok{\# But a PROTECTED document with a missing classification field}
    \CommentTok{\# is now labelled OFFICIAL and treated accordingly.}
    \ControlFlowTok{return}\NormalTok{ classification}
\end{Highlighting}
\end{Shaded}

The three versions illustrate a spectrum. The offensive version turns a
crash into an actionable incident --- the operator knows which document,
what data was present, and what to investigate. The bare access version
at least crashes, which is correct behaviour for a data integrity
failure, but the operator gets a generic \texttt{KeyError} with no
diagnostic context. The agent-generated version is the worst outcome: it
doesn't crash at all, silently fabricating a classification that
downstream access control decisions will treat as authoritative.

The agent-generated version:

\begin{itemize}
\tightlist
\item
  Is syntactically valid Python
\item
  Passes every unit test (the default prevents even downstream and
  integration tests from detecting a problem)
\item
  Follows the \texttt{.get()} pattern that appears in millions of Python
  files in the agent's training data
\item
  Would pass casual code review (it looks like defensive, robust coding)
\item
  Silently downgrades document classifications when data integrity
  failures occur
\end{itemize}

While the example is ficticious, this is the pattern that defensive
programming produces \emph{by default} in Python, and agents are trained
on defensive Python.

\hypertarget{what-is-fundamentally-different-about-agentic-code}{%
\subsection{2.4 What Is Fundamentally Different About Agentic
Code}\label{what-is-fundamentally-different-about-agentic-code}}

The threat model for agent-generated code is not simply ``human-authored
code but more of it.'' Several properties are qualitatively different:

\textbf{Limited persistent learning.} A human developer who receives
review feedback on a trust boundary violation learns from it and is less
likely to repeat the mistake. Agents have limited or no persistent
memory across sessions. The practical consequence is stark: the agent
isn't circumventing project rules, and it isn't ignoring instructions
--- it followed them perfectly in the last session. It just doesn't
\emph{have} a last session. Every invocation is the first day on the
job, and on the first day, you write \texttt{.get()} with a default
because that's what Python looks like in the training data.

Some agent frameworks now support project-level instructions (system
prompts, documentation files, memory stores) that provide partial
mitigation. An agent can be told ``don't use \texttt{.get()} on audit
data'' and will follow that instruction within a session. But these are
explicit rules, not internalised judgment. The agent cannot generalise
from ``don't use \texttt{.get()} on audit data'' to ``don't fabricate
defaults anywhere that data absence is meaningful'' unless that
generalisation is also spelled out. Every correction must be encoded as
a rule; the agent does not learn the \emph{principle} behind the
correction. This means that \emph{review feedback improves the generator
only to the extent that it is captured as machine-readable rules} ---
and the coverage of those rules is always trailing the set of possible
failure modes.

\textbf{Consistent surface quality.} Human code has variable surface
quality --- hasty code looks hasty, careful code looks careful.
Reviewers use surface quality as a signal for where to focus attention.
Agent code has uniformly high surface quality regardless of semantic
correctness. A function with a critical trust boundary violation looks
exactly as polished as a function without one. The reviewer's natural
calibration signal --- ``this code looks sloppy, I should look more
carefully'' --- is absent.

\textbf{Pattern completion, not intent.} A human developer writing
\texttt{record.get("classification",\ "OFFICIAL")} has either made a
deliberate design decision (the default is intentional) or made an error
(they didn't think about the missing-field case). The distinction is
visible in context --- comments, commit messages, design docs. An agent
writing the same code is completing a pattern from training data. It has
no design intent. There is no commit message that explains why the
default is correct, because the agent didn't decide it was correct ---
it predicted it was the next likely token.

\textbf{Intent-based review (``why did you write it this way?'') is
meaningless for agent code.} The review must be entirely outcome-based
(``is the behaviour correct for this context?'').

\textbf{Correlated failure modes.} When ten human developers write code
for a system, their errors are largely independent --- different people
make different mistakes. When an agent generates ten functions, its
errors are \emph{correlated} --- the same training data biases produce
the same failure modes repeatedly. A single systematic bias (e.g.,
``always use \texttt{.get()} with a default'') produces correlated
vulnerabilities across the entire codebase. This is not the
independent-error model that code review and testing strategies are
designed for.

\textbf{No fatigue, no shortcuts --- but also no judgment.} Agents don't
get tired, don't take shortcuts under deadline pressure, and don't
introduce bugs from distraction. But they also don't exercise judgment
about which patterns are appropriate in which contexts. A human
developer who is tired might introduce a bug in one function; an agent
that lacks context will introduce the same incorrect pattern in every
function it generates. The failure mode is not degradation under
pressure --- it is \emph{systematic misapplication of
context-inappropriate patterns}.

\textbf{Model monoculture amplifies correlation across organisational
boundaries.} The correlated failure problem described above operates
within a single codebase, but it extends further. If 80\% of government
agencies adopt the same two or three models for code generation, the
correlated failure modes are no longer contained within individual
organisations. A systematic bias in a widely-used model --- say, a
persistent tendency to use \texttt{.get()} with defaults on
security-critical fields --- will produce the same vulnerability pattern
across every codebase that model touches. This is analogous to
agricultural monoculture: genetic uniformity makes the entire crop
vulnerable to a single pathogen. Discovering a systematic
agent-introduced defect pattern in one agency should trigger
cross-agency scanning, because the same model likely introduced the same
pattern elsewhere. This strengthens the case for cross-organisational
standards (Section 9.6) and shared vulnerability disclosure mechanisms
for agent-introduced defect patterns.

\hypertarget{why-training-data-is-the-root-cause}{%
\subsection{2.5 Why Training Data Is the Root
Cause}\label{why-training-data-is-the-root-cause}}

The vast majority of open-source Python code uses defensive patterns:
\texttt{.get()} with defaults, \texttt{getattr()} with fallbacks, broad
\texttt{try/except} blocks, \texttt{or} chains that silently substitute
values. These patterns are appropriate for applications where graceful
degradation is preferable to crashing --- which is most applications.

They are catastrophically inappropriate for applications where:

\begin{itemize}
\tightlist
\item
  Silent data corruption is worse than a crash
\item
  Every decision must be traceable to its source
\item
  The absence of data is itself evidence (not an invitation to fabricate
  a default)
\item
  Error paths must be as auditable as success paths
\end{itemize}

Government systems handling classified information, financial records,
health data, or law enforcement evidence fall squarely into this
category. The agent doesn't know this. It can't know this from the code
alone --- the security context is institutional knowledge, not syntactic
structure.

\hypertarget{stride-applied-to-agentic-code-output}{%
\section{3. STRIDE Applied to Agentic Code
Output}\label{stride-applied-to-agentic-code-output}}

\hypertarget{framework-selection}{%
\subsection{3.1 Framework Selection}\label{framework-selection}}

STRIDE (Spoofing, Tampering, Repudiation, Information Disclosure, Denial
of Service, Elevation of Privilege) is the established threat modelling
framework used in Australian government security assessments. Applying
it to agentic code output provides a structured vocabulary that policy
audiences already understand, rather than inventing new terminology.

The application below extends STRIDE to treat \textbf{agent-generated
code as an input} to the system --- analogous to treating user input as
untrusted. The agent is not an adversary, but its output has the same
trust properties as any external input: it may be well-formed, it may be
reasonable, but it has not been validated against the system's security
requirements. Two of the six categories below --- ``competence
spoofing'' (S) and the process-level DoS entries (D) --- are
\textbf{analogical extensions} of STRIDE's original technical-system
categories to development-process analysis, not standard applications.
STRIDE-LM (adding Lateral Movement; see e.g., Meier et al.~on CSF Tools)
and the emerging ASTRIDE variant (adding AI Agent-Specific Attacks;
Dutta et al.~2025) provide precedent for such extensions; nonetheless,
the reader should understand these as novel mappings rather than
established STRIDE doctrine.

\textbf{A note on taxonomy levels.} The threat categories below
deliberately mix code-level semantic failures (S, T, I, E),
process-level failures (D), and assurance degradation. This mixing is
intentional, not accidental. The threat model's central argument is that
these levels \emph{interact}: code-level failures (e.g., trust tier
conflation) pass review \emph{because} of process-level failures
(finding flood, review capacity exhaustion). Separating them into clean
categories would be more formally tidy but would obscure the compounding
effect (Section 3.3) that makes the threat model novel. Where an entry
is a process threat rather than a code pattern, it is marked as such.

\hypertarget{threat-categories}{%
\subsection{3.2 Threat Categories}\label{threat-categories}}

\hypertarget{s-spoofing-competence-spoofing}{%
\subsubsection{S --- Spoofing: Competence
Spoofing}\label{s-spoofing-competence-spoofing}}

\textbf{Traditional STRIDE scope:} An entity claims to be something it
is not --- a forged authentication token, a spoofed IP address, a
process impersonating another user. The system accepts the false
identity and grants access or trust accordingly.

\textbf{Agentic variant:} Code \emph{appears} to handle data correctly
but operates on fabricated or default values, presenting a false picture
of data integrity.

\textbf{Mechanism:} Agents default to defensive patterns that substitute
values rather than failing. The code ``works'' --- it produces output,
it doesn't crash --- but the output is based on fabricated data rather
than actual data. The code spoofs the competence of correct data
handling.

\textbf{Examples:}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# Fabricates a default rather than surfacing missing data}
\NormalTok{user\_role }\OperatorTok{=} \BuiltInTok{getattr}\NormalTok{(session, }\StringTok{"role"}\NormalTok{, }\StringTok{"readonly"}\NormalTok{)}
\CommentTok{\# If session.role doesn\textquotesingle{}t exist (bug, tampering, schema change),}
\CommentTok{\# user silently gets "readonly" — which might be WRONG in either direction.}
\CommentTok{\# If they should have been "admin", they can\textquotesingle{}t work.}
\CommentTok{\# If "readonly" still has access to sensitive data, it\textquotesingle{}s a privilege grant.}

\CommentTok{\# Presents fabricated confidence}
\NormalTok{confidence }\OperatorTok{=}\NormalTok{ result.get(}\StringTok{"confidence\_score"}\NormalTok{, }\FloatTok{0.5}\NormalTok{)}
\CommentTok{\# A missing confidence score defaults to "medium confidence"}
\CommentTok{\# rather than "unknown" — downstream decisions treat this as real data.}

\CommentTok{\# Fabricates identity via structural presence}
\ControlFlowTok{if} \BuiltInTok{hasattr}\NormalTok{(obj, }\StringTok{"security\_clearance"}\NormalTok{):}
\NormalTok{    handle\_classified(obj)}
\CommentTok{\# Any object that declares this attribute passes the gate.}
\CommentTok{\# The check is structural (has the attribute), not ontological (is the type).}
\CommentTok{\# A trivial Impersonator class with security\_clearance = "TOP\_SECRET" gets in.}
\end{Highlighting}
\end{Shaded}

\textbf{Why existing controls miss it:} The code is syntactically valid,
follows common patterns, and passes tests. A human reviewer under time
pressure sees ``defensive coding'' --- a positive signal. The
fabrication is invisible without understanding the security semantics of
each field.

\textbf{Risk in government context:} Classification decisions, access
control, evidentiary integrity --- any domain where ``I don't know'' and
``the default'' are different answers with different consequences.

\hypertarget{t-tampering-silent-trust-tier-coercion}{%
\subsubsection{T --- Tampering: Silent Trust Tier
Coercion}\label{t-tampering-silent-trust-tier-coercion}}

\textbf{Traditional STRIDE scope:} An attacker modifies data in transit
or at rest without authorisation --- altering a database record,
intercepting and changing a message, corrupting a configuration file.
The data itself is changed.

\textbf{Agentic variant:} External (untrusted) data is treated as
internal (trusted) data without validation, effectively tampering with
the trust level rather than the data itself.

\textbf{Mechanism:} Agents don't distinguish between data from different
trust levels because the programming language doesn't enforce it. A
\texttt{dict} from a validated database query and a \texttt{dict} from
an unvalidated API response are the same type. The agent treats them
interchangeably.

\textbf{Examples:}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# API response used directly without validation boundary}
\NormalTok{api\_response }\OperatorTok{=}\NormalTok{ requests.get(external\_url).json()}
\NormalTok{save\_to\_internal\_database(api\_response[}\StringTok{"records"}\NormalTok{])}
\CommentTok{\# External data enters the trusted internal store without validation.}
\CommentTok{\# The data might be malformed, might contain injection payloads,}
\CommentTok{\# might be missing required fields.}
\CommentTok{\# The agent doesn\textquotesingle{}t see a trust boundary — it sees a dict going into a function.}

\CommentTok{\# Deserialized data assumed trustworthy}
\NormalTok{config }\OperatorTok{=}\NormalTok{ json.loads(uploaded\_config\_file.read())}
\NormalTok{apply\_system\_settings(config)}
\CommentTok{\# User{-}uploaded JSON treated as trusted configuration.}
\CommentTok{\# No schema validation, no field allowlisting, no type checking.}
\end{Highlighting}
\end{Shaded}

\textbf{Why existing controls miss it:} Type checkers verify shape
(\texttt{dict}), not provenance (where the dict came from). Linters
check syntax, not data flow. The code has no detectable defect at any
analysis level short of semantic boundary tracking.

\textbf{Risk in government context:} Injection attacks through
unvalidated external data, data corruption of authoritative records,
compliance failures when data provenance cannot be demonstrated.

\hypertarget{r-repudiation-audit-trail-destruction-through-error-handling}{%
\subsubsection{R --- Repudiation: Audit Trail Destruction Through Error
Handling}\label{r-repudiation-audit-trail-destruction-through-error-handling}}

\textbf{Traditional STRIDE scope:} A user performs an action --- a
transaction, an access, a deletion --- and later denies it. The system
lacks sufficient logging, signing, or non-repudiation controls to prove
the action occurred. The gap is in the \emph{recording} of events.

\textbf{Agentic variant:} Error handling patterns destroy the audit
trail by catching, logging, and continuing rather than failing in a way
that preserves the error as a first-class audit event.

\textbf{Mechanism:} Agents generate broad exception handlers that
prevent crashes but also prevent errors from being recorded in audit
systems. The error is ``handled'' in the sense that the program
continues, but the event that caused the error is lost to the audit
trail.

\textbf{Examples:}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# Error swallowed — audit trail has a gap}
\ControlFlowTok{try}\NormalTok{:}
\NormalTok{    record\_decision(case\_id, decision, evidence)}
\ControlFlowTok{except} \PreprocessorTok{Exception} \ImportTok{as}\NormalTok{ e:}
\NormalTok{    logger.warning(}\SpecialStringTok{f"Failed to record decision: }\SpecialCharTok{\{}\NormalTok{e}\SpecialCharTok{\}}\SpecialStringTok{"}\NormalTok{)}
    \CommentTok{\# Decision was made but not recorded.}
    \CommentTok{\# If this is a legal/regulatory decision, the audit trail}
    \CommentTok{\# now has a gap that cannot be reconstructed.}
    \CommentTok{\# The log message may be rotated, compressed, or lost.}
    \CommentTok{\# The audit database — the authoritative record — shows nothing.}

\CommentTok{\# Partial completion without rollback}
\ControlFlowTok{try}\NormalTok{:}
\NormalTok{    update\_classification(document\_id, new\_level)}
\NormalTok{    notify\_stakeholders(document\_id, new\_level)}
\NormalTok{    record\_classification\_change(document\_id, old\_level, new\_level)}
\ControlFlowTok{except}\NormalTok{ NotificationError:}
    \ControlFlowTok{pass}  \CommentTok{\# "Notification is non{-}critical"}
    \CommentTok{\# Classification changed, stakeholders not notified, change not recorded.}
    \CommentTok{\# Three operations that should be atomic are silently partial.}
\end{Highlighting}
\end{Shaded}

\textbf{Why existing controls miss it:} The code handles exceptions ---
which is generally considered good practice. The distinction between
``handle the error and continue safely'' and ``swallow the error and
destroy evidence'' requires understanding which operations are
audit-critical. The agent doesn't know.

\textbf{Risk in government context:} Regulatory compliance (failure to
maintain complete audit trails), legal proceedings (gaps in evidence
chains), IRAP assessment failures (inability to demonstrate complete
traceability).

\hypertarget{i-information-disclosure-verbose-error-responses}{%
\subsubsection{I --- Information Disclosure: Verbose Error
Responses}\label{i-information-disclosure-verbose-error-responses}}

\textbf{Traditional STRIDE scope:} Sensitive data is exposed to
unauthorised parties --- through a misconfigured access control, a
side-channel leak, an unprotected API endpoint, or data at rest without
encryption. The exposure is typically of \emph{stored or transmitted}
data.

\textbf{Agentic variant:} Agent-generated error handling exposes
internal system details in error responses, log messages, or API
returns.

\textbf{Mechanism:} Agents produce ``helpful'' error messages that
include internal state, query parameters, file paths, or stack traces.
This is good practice for development but dangerous in production, and
agents don't distinguish between the two contexts.

\textbf{Examples:}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# Agent{-}generated "helpful" error handler}
\ControlFlowTok{except}\NormalTok{ DatabaseError }\ImportTok{as}\NormalTok{ e:}
    \ControlFlowTok{return}\NormalTok{ \{}
        \StringTok{"error"}\NormalTok{: }\BuiltInTok{str}\NormalTok{(e),}
        \StringTok{"query"}\NormalTok{: sql\_query,         }\CommentTok{\# Exposes database schema}
        \StringTok{"connection"}\NormalTok{: }\BuiltInTok{str}\NormalTok{(db\_url),  }\CommentTok{\# May contain credentials}
        \StringTok{"params"}\NormalTok{: query\_params,     }\CommentTok{\# Exposes internal identifiers}
\NormalTok{    \}}

\CommentTok{\# Stack trace in API response}
\ControlFlowTok{except} \PreprocessorTok{Exception} \ImportTok{as}\NormalTok{ e:}
    \ImportTok{import}\NormalTok{ traceback}
    \ControlFlowTok{return}\NormalTok{ \{}\StringTok{"error"}\NormalTok{: traceback.format\_exc()\}}
    \CommentTok{\# Full stack trace including file paths, function names,}
    \CommentTok{\# library versions — a reconnaissance goldmine.}
\end{Highlighting}
\end{Shaded}

\textbf{Why existing controls miss it:} The error handling is
syntactically correct and genuinely helpful during development.
Detecting that internal details should not appear in production error
responses requires understanding the deployment context, not the code
structure.

\textbf{Risk in government context:} Reconnaissance information for
attackers, credential exposure, violation of need-to-know principles.

\hypertarget{d-denial-of-service-finding-flood-meta-threat}{%
\subsubsection{D --- Denial of Service: Finding Flood
(Meta-Threat)}\label{d-denial-of-service-finding-flood-meta-threat}}

\textbf{Traditional STRIDE scope:} An attacker exhausts a system's
resources --- network bandwidth, CPU, memory, connection pools ---
rendering it unable to serve legitimate requests. The attack targets
\emph{availability} of a runtime service.

\textbf{Agentic variant:} The volume of agent-generated code overwhelms
the review process, degrading review quality to the point where the
review is no longer an effective security control.

\emph{Note: This is an extended application of STRIDE to the development
lifecycle, not the runtime system. The ``service'' being denied is the
review process --- a security control, not a user-facing system. This
extension is deliberate: if the review process is a security control
(and it is, per ISM-2060/2061), then degrading that control's
effectiveness is a denial-of-service against the security posture, even
though no runtime system is affected.}

\textbf{Mechanism:} This is not a code pattern --- it is a
\emph{process} threat. When agents generate code at multiples of human
velocity, the review queue grows proportionally. Reviewers under volume
pressure shift from careful semantic review to surface-level scanning.
The review process --- which is a security control --- degrades to a
rubber stamp.

A secondary mechanism: when automated analysis tools produce too many
findings on agent-generated code, reviewers habituate to dismissing
findings, and genuine security issues are lost in the noise.

\textbf{Why existing controls miss it:} Existing controls assume review
capacity scales with code generation rate. It doesn't. The control's
effectiveness is inversely proportional to the volume it processes,
which is the opposite of every other scaling assumption in the process.

\textbf{Risk in government context:} Security review as a compliance
checkbox rather than an effective control, accreditation based on a
process that no longer provides the assurance it claims to provide.

\hypertarget{e-elevation-of-privilege-trust-tier-conflation}{%
\subsubsection{E --- Elevation of Privilege: Trust Tier
Conflation}\label{e-elevation-of-privilege-trust-tier-conflation}}

\textbf{Traditional STRIDE scope:} A user or process gains capabilities
beyond what is authorised --- exploiting a kernel vulnerability to move
from user to root, leveraging a misconfigured role to access admin
functions, or escaping a sandbox to reach the host system. The
escalation is of \emph{identity or access rights}.

\textbf{Agentic variant:} Data from a lower trust level is used in a
higher-trust context without explicit validation, effectively elevating
the data's privilege level.

\textbf{Mechanism:} Closely related to Tampering (Section 3.2, T), but
focussed on the \emph{consequence} rather than the \emph{mechanism}.
When external data enters internal processing without validation, any
actions taken based on that data inherit a trust level they haven't
earned.

\textbf{Examples:}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# User{-}supplied filter used in internal query without validation}
\KeywordTok{def}\NormalTok{ search\_records(user\_query: }\BuiltInTok{dict}\NormalTok{):}
    \CommentTok{\# user\_query comes from an API request (untrusted)}
\NormalTok{    results }\OperatorTok{=}\NormalTok{ db.query(Record).filter\_by(}\OperatorTok{**}\NormalTok{user\_query)}
    \CommentTok{\# **user\_query unpacks arbitrary keys into the query filter.}
    \CommentTok{\# User can filter on internal fields (is\_deleted, internal\_score)}
    \CommentTok{\# that should not be exposed.}
    \ControlFlowTok{return}\NormalTok{ results}

\CommentTok{\# External system\textquotesingle{}s assertion accepted without verification}
\NormalTok{partner\_response }\OperatorTok{=}\NormalTok{ partner\_api.verify\_identity(applicant\_id)}
\ControlFlowTok{if}\NormalTok{ partner\_response.get(}\StringTok{"verified"}\NormalTok{, }\VariableTok{False}\NormalTok{):}
\NormalTok{    grant\_access(applicant\_id)}
    \CommentTok{\# Partner system\textquotesingle{}s assertion of identity is accepted at face value.}
    \CommentTok{\# No independent verification, no recording of the basis for the decision.}
    \CommentTok{\# The partner\textquotesingle{}s trust level is silently elevated to our trust level.}
\end{Highlighting}
\end{Shaded}

\textbf{Why existing controls miss it:} The code follows common
integration patterns. The privilege elevation is implicit --- no
explicit \texttt{setRole()} or \texttt{grantPermission()} call that a
scanner could detect. The elevation happens through data flow, not code
structure.

\textbf{Risk in government context:} Unauthorized access to classified
information, acceptance of unverified identity assertions in federated
systems, compliance failures in inter-agency data sharing.

\hypertarget{the-compounding-effect}{%
\subsection{3.3 The Compounding Effect}\label{the-compounding-effect}}

These six threat categories do not operate independently. In practice,
they compound. One illustrative scenario:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  An agent generates code with \textbf{trust tier conflation} (E) ---
  external API data used directly
\item
  The missing validation means errors in that data are caught by a broad
  \texttt{except} block --- \textbf{audit trail destruction} (R)
\item
  The \texttt{except} block returns a default value ---
  \textbf{competence spoofing} (S)
\item
  The default value is treated as authoritative data downstream ---
  \textbf{silent trust tier coercion} (T)
\item
  The volume of agent-generated code means the reviewer doesn't catch
  any of this --- \textbf{finding flood} (D)
\end{enumerate}

Each individual pattern looks like ordinary defensive coding. The
compound effect is a system that silently produces wrong results, can't
explain why, and passed every review gate.

\hypertarget{the-review-process-as-attack-surface}{%
\section{4. The Review Process as Attack
Surface}\label{the-review-process-as-attack-surface}}

\hypertarget{the-asymmetry-problem}{%
\subsection{4.1 The Asymmetry Problem}\label{the-asymmetry-problem}}

Human code review evolved as a control for human-authored code at human
pace. It relies on several assumptions that agentic coding violates:

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.2619}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.3571}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.3810}}@{}}
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Assumption
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Human Code
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Agent Code
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
\textbf{Volume} & Reviewers can read most of the code & Volume makes
comprehensive reading impossible \\
\textbf{Familiarity} & The author can explain their intent & The agent's
``intent'' is pattern completion from training data \\
\textbf{Conventions} & Unusual patterns are suspicious & Agents follow
conventions precisely --- the dangerous code looks normal \\
\textbf{Error rate} & Human error rate is \textasciitilde1-5\% per
function & Agent error rate for \emph{semantic} correctness is unknown
and context-dependent \\
\textbf{Feedback} & Reviewer feedback improves the author & Agent has no
persistent memory across review cycles \\
\end{longtable}

The consequences of this asymmetry are already visible at scale.
GitHub's February 2026 announcement of platform-level PR restrictions
(Section 1.2.2) is a direct institutional response to the volume
assumption breaking down in the open-source ecosystem.

\hypertarget{the-habituation-effect}{%
\subsection{4.2 The Habituation Effect}\label{the-habituation-effect}}

When agents generate code that consistently passes tests and follows
conventions, reviewers develop trust in the agent's output. This trust
is not earned --- it is a cognitive shortcut driven by volume pressure.
In human factors engineering, this phenomenon is known as
\textbf{automation bias}: the tendency to over-rely on automated systems
and under-scrutinise their output. The effect is well-documented in
aviation safety, medical decision support, and industrial automation
(Parasuraman \& Manzey 2010) --- and recent research demonstrates it
manifests specifically in AI-assisted software development. Perry et
al.~(2023) found that developers with AI coding access wrote less secure
code while simultaneously \emph{feeling more confident} about their
code's security --- a textbook automation bias outcome. The METR
randomised controlled trial (2025) found experienced developers
predicted AI would speed them up by 24\%, believed after using it that
they were 20\% faster, but were measured as 19\% \emph{slower} --- a
43-percentage-point perception-reality gap that demonstrates the
phenomenon operating in real-world coding conditions.

The reviewer's mental model shifts from ``verify this code is correct''
to ``check this code isn't obviously wrong.'' The difference is
enormous: the first is an active search for defects; the second is a
passive scan that catches only gross errors.

The habituation effect described above has been directly observed in
practice. In compliance-constrained development environments,
agent-generated code containing semantic defects --- trust boundary
violations, defensive patterns on audit-critical data, missing
validation boundaries --- has entered codebases after passing human
review processes that were designed to catch exactly these issues. The
defects were subsequently identified through other means: automated
enforcement tooling, adjacent code review, or downstream test failures.
The pattern is consistent with the automation bias literature: the
code's surface quality satisfied the reviewer's heuristic threshold, and
the semantic violation was not visible at the inspection depth that
volume pressure permits.

This is the ``Shifting the Burden'' systems archetype (Meadows 2008):
the agent's consistent surface-quality output becomes the symptomatic
fix that weakens the fundamental solution (thorough human review). The
more the agent produces acceptable-looking code, the less carefully
humans review it, and the more dependent the process becomes on the
agent being correct --- which is exactly the assumption the review
process exists to check.

A related but distinct mechanism compounds this effect. Agent-assisted
velocity increases the \emph{parallelisation} of work, not just its
speed. When an agent assists in producing multiple interdependent
artefacts simultaneously --- a design specification, an implementation,
and a policy document --- semantic inconsistencies \emph{between}
artefacts become invisible because no single review pass covers all of
them. The review window for cross-document consistency shrinks in
proportion to the velocity gain. The reviewer is not only less careful
per artefact, but also unable to hold the full production context in
working memory at the rate artefacts are produced.

The availability of parallel agent generation creates a structural
pressure that procedural and behavioural controls may mitigate but are
unlikely to eliminate, because the same incentives driving adoption also
reward bypassing throughput-constraining review practices. An
organisation can prohibit developers from running multiple agents
concurrently, but the prohibition runs directly against the productivity
incentive that justified adopting agentic development. Controls that
depend on sustained human restraint in the face of convenience are
inherently fragile --- a principle well-established in security
engineering but easy to overlook when the convenience is ``generate code
faster than you can review it.'' The implications for control selection
are examined in Section 7.

\hypertarget{the-advisory-fatigue-problem}{%
\subsection{4.3 The Advisory Fatigue
Problem}\label{the-advisory-fatigue-problem}}

Static analysis tools that flag agent-generated code patterns as
warnings face a paradox:

\begin{itemize}
\tightlist
\item
  If agents produce many warnings, reviewers habituate to dismissing
  them
\item
  If agents learn to avoid warning-triggering patterns, they may produce
  code that satisfies the tool but still violates the semantic intent
\item
  Advisory-only tools have no enforcement mechanism for agents, which
  have no memory across sessions --- a warning shown to an agent in one
  session is forgotten in the next
\end{itemize}

This means the traditional ``warn first, enforce later'' adoption
strategy for security tooling is ineffective for agent-generated code.
Agents require enforcement at the boundary (before code enters the
repository), not feedback over time (which requires learning).

\hypertarget{agent-output-as-a-trust-boundary}{%
\section{5. Agent Output as a Trust
Boundary}\label{agent-output-as-a-trust-boundary}}

\hypertarget{the-trust-tier-model}{%
\subsection{5.1 The Trust Tier Model}\label{the-trust-tier-model}}

Data in high-assurance systems can be classified by provenance into
trust tiers:

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.1395}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.3023}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.3488}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2093}}@{}}
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Tier
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Description
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Handling Rule
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Example
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
\textbf{Tier 1: Internal data} & Data authored by the system itself ---
audit records, internal state, configuration & Full trust. Any anomaly
indicates corruption or tampering --- fail immediately, do not attempt
recovery & Database audit trail, system configuration, internal state
machines \\
\textbf{Tier 2: Validated data} & Data that entered the system from
outside but has been validated at the boundary & Elevated trust. Types
are reliable; values may still cause operational failures & API response
after schema validation, CSV row after type coercion \\
\textbf{Tier 3: External data} & Data from outside the system boundary,
not yet validated & Zero trust. May be malformed, malicious, or missing.
Validate at boundary, quarantine failures & Raw API responses, user
uploads, message queue payloads \\
\end{longtable}

This model is standard practice for data. External inputs are validated
at the perimeter because the cost and risk of allowing unvalidated data
to traverse internal systems is not worth the convenience of deferring
the check.

\hypertarget{agent-code-as-tier-3}{%
\subsection{5.2 Agent Code as Tier 3}\label{agent-code-as-tier-3}}

The central proposal of this paper applies the same principle to code:
\textbf{agent-generated code should be treated as Tier 3 (external,
untrusted) data until validated.}

This is not a statement about agent quality --- agents produce excellent
code much of the time. It is a statement about \emph{provenance}. The
agent is an external system. Its output has not been validated against
the system's security requirements. The fact that the output is source
code rather than JSON or CSV does not change its trust properties --- it
warrants the same boundary discipline.

Treating agent code as Tier 3 has specific implications:

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.4783}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.5217}}@{}}
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Principle
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Application
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
\textbf{Validate at the boundary} & Agent output must pass
security-aware validation before entering the codebase \\
\textbf{Quarantine failures} & Code that fails validation is rejected,
not silently corrected \\
\textbf{Record what we got} & The original agent output is preserved for
audit, even if modified during review \\
\textbf{No silent coercion} & Agent code is not silently ``fixed up'' by
reviewers --- changes are explicit and recorded \\
\end{longtable}

\hypertarget{implications-for-the-development-workflow}{%
\subsection{5.3 Implications for the Development
Workflow}\label{implications-for-the-development-workflow}}

If agent output is Tier 3, the development workflow must include a
\textbf{validation boundary} between agent generation and code
integration:

\begin{center}
\begin{BVerbatim}
Agent generates code
        │
        v
┌──────────────────────┐
│  VALIDATION BOUNDARY │  ← This is the trust boundary
│                      │
│  • Automated semantic│     Not just syntax/type checking —
│    boundary checking │     trust tier flow, defensive pattern
│  • Human review of   │     detection, audit trail completeness
│    semantic intent   │
│  • Attestation       │     Reviewer attests validation was
│                      │     meaningful, not rubber-stamped
└──────────────────────┘
        │
        v
Code enters repository
(now Tier 2 — validated)
\end{BVerbatim}
\end{center}

The key difference from current practice: \textbf{the validation is
security-aware, not just correctness-aware.} Current code review asks
``does this code work?'' Security-aware validation asks ``does this code
maintain the system's trust boundaries?''

\hypertarget{current-guidance-gap-analysis}{%
\section{6. Current Guidance Gap
Analysis}\label{current-guidance-gap-analysis}}

\hypertarget{australian-information-security-manual-ism}{%
\subsection{6.1 Australian Information Security Manual
(ISM)}\label{australian-information-security-manual-ism}}

The ISM provides controls for software development security (primarily
in the Software Development and Web Application Development chapters).
The analysis below maps relevant controls to the agentic code threat
model, identifying where existing controls provide partial coverage,
where they assume conditions that agentic coding invalidates, and where
gaps exist.

\emph{Note: The ISM underwent a significant expansion in June 2025,
adding approximately 24 new controls to the Software Development
guidelines. The analysis below references the December 2025 revision of
the ISM. Organisations using earlier versions should verify control
numbers against the current release.}

\hypertarget{controls-with-partial-coverage}{%
\subsubsection{6.1.1 Controls with Partial
Coverage}\label{controls-with-partial-coverage}}

The following controls provide partial coverage of agentic threats. For
each, we identify what the control currently addresses and where its
assumptions break down when applied to agent-generated code.

\textbf{ISM-0401} (Rev 8, Jun-25) --- \emph{Secure by Design principles
and practices throughout the SDLC}

\emph{Coverage:} Establishes that organisations should follow Secure by
Design principles across the entire software development lifecycle.
Agentic failure modes (Appendix A) could in principle be addressed as
part of an organisation's Secure by Design practices.

\emph{Gap:} The control assumes a human development team that can
\emph{internalise} security principles and apply them with judgment.
Agents don't internalise principles --- they reproduce training data
patterns. A Secure by Design practice that says ``don't fabricate
defaults for missing safety-critical data'' is unenforceable against an
agent unless encoded as a machine-checkable rule. The control's scope
(whole SDLC) is correct, but its enforcement mechanism (human judgment)
doesn't transfer to agent-generated code.

\textbf{ISM-1419} (Rev 1, Sep-18) --- \emph{Development and modification
of software only in development environments}

\emph{Coverage:} Requires segregation of development from operational
environments. This is orthogonal to agentic threats --- it constrains
\emph{where} code is written, not \emph{how} or \emph{by whom}.

\emph{Gap:} The control provides no coverage of agentic code quality or
review. Its value is environmental separation, which remains important
(agents should not have direct access to production environments) but
does not address the semantic correctness of agent-generated code.

\textbf{ISM-2060} (Rev 0, Jun-25) --- \emph{Code reviews ensure software
meets Secure by Design principles and secure programming practices}

\emph{Coverage:} Directly applicable to agent-generated code --- the
agent is ``the author'' and a human is ``the reviewer.'' The control
explicitly links code review to Secure by Design principles, not just
functional correctness.

\emph{Gap:} The control assumes the reviewer can meaningfully evaluate
the code at the rate it is produced. At agent-scale volume, this
assumption fails (Section 4). The control does not address review
effectiveness degradation, nor does it distinguish between surface-level
review (syntax, conventions) and security-focused review (trust
boundaries, audit trail integrity).

\textbf{ISM-2061} (Rev 0, Jun-25) --- \emph{Security-focused peer
reviews on critical and security-focused software components}

\emph{Coverage:} Requires developer-supported, security-focused peer
reviews specifically on critical components. This is the strongest
existing review control for the agentic context.

\emph{Gap:} The control's limitation is scope: it applies to ``critical
and security-focused software components,'' which requires the
organisation to correctly identify which agent-generated code touches
security-critical paths. Agents generate code across the entire
codebase; the security-critical subset must be identified before the
review control can be applied. The control also assumes the peer
reviewer has the institutional knowledge to evaluate trust boundary
maintenance --- knowledge that may not be documented in machine-readable
form.

\textbf{ISM-0402} (Rev 9, Jun-25) --- \emph{Comprehensive software
testing using SAST, DAST, and SCA}

\emph{Coverage:} Mandates static application security testing (SAST),
dynamic application security testing (DAST), and software composition
analysis (SCA). These tools catch known vulnerability patterns and
dependency risks.

\emph{Gap:} The failure modes in this threat model are specifically
designed to pass existing SAST/DAST tools (Section 2.3). Current SAST
catches ``does the code contain known vulnerability patterns?'' but not
``does the code maintain trust boundaries it doesn't know about?''
Semantic boundary testing --- verifying that data flows respect trust
tiers --- is a distinct category not addressed by existing SAST tooling.
SCA is relevant for agent-introduced dependencies but does not address
first-party code quality.

\textbf{ISM-2026/2027/2028} (Jun-25) --- \emph{Software artefact
integrity --- malicious code scanning, digital signatures, SAST/DAST/SCA
on artefacts}

\emph{Coverage:} Addresses integrity and security scanning of software
artefacts before deployment. These controls cover the supply chain from
build to deployment.

\emph{Gap:} Agent-generated first-party code is a novel supply chain
input --- it's code that appears in-house but was produced by an
external system (the AI model). The artefact integrity controls don't
have a category for ``first-party code generated by a third-party
system.'' The risk properties are also different: third-party components
have independent defect distributions, while agent-generated code has
correlated defects (Section 2.4). The controls verify artefact integrity
but not the semantic correctness of the code within those artefacts.

\hypertarget{controls-with-no-coverage}{%
\subsubsection{6.1.2 Controls with No
Coverage}\label{controls-with-no-coverage}}

The following gap areas have no corresponding ISM control. Each
represents a category of agentic risk that falls outside the current
framework's scope.

\textbf{Agent output as trust boundary} --- \emph{ACF-T1 (trust tier
conflation), ACF-E1 (implicit privilege grant)}

No control addresses the trust classification of AI-generated artefacts.
Agent code is neither ``in-house'' (human-authored) nor ``third-party''
(external component) --- it's a new category. ISM-2074 (Dec-25) requires
organisations to develop, implement, and maintain an AI usage policy,
but this is a governance control, not a technical trust boundary
control.

\textbf{Review capacity scaling} --- \emph{ACF-D1 (finding flood),
ACF-D2 (review capacity exhaustion)}

ISM-2060 and ISM-2061 mandate code review and security-focused peer
review, but neither addresses what happens when code generation velocity
exceeds review capacity. No control requires organisations to
demonstrate that review remains effective under volume pressure.

\textbf{Semantic boundary enforcement} --- \emph{ACF-S1 (competence
spoofing), ACF-S3 (structural identity spoofing), ACF-T1 (trust tier
conflation), ACF-T2 (silent coercion)}

No control addresses the gap between syntactic correctness and semantic
correctness in the context of trust boundaries. ISM-0402's SAST/DAST/SCA
requirement covers known vulnerability patterns but not
context-dependent semantic correctness. Existing controls assume that if
code passes review and testing, it is adequate.

\textbf{Correlated failure detection} --- \emph{All ACF categories}

No control addresses the distinct risk profile of correlated defects.
Testing and review strategies are designed for independent failure
distributions.

\textbf{Code provenance tracking} --- \emph{ACF-D2 (review capacity
exhaustion)}

No control requires organisations to track which code was generated by
AI agents vs.~authored by humans. ISM-2074 requires an AI usage policy
but not per-artefact provenance. Without provenance, risk assessment
cannot distinguish between code populations with different failure
characteristics.

\hypertarget{candidate-ism-extensions}{%
\subsubsection{6.1.3 Candidate ISM
Extensions}\label{candidate-ism-extensions}}

The following are illustrative extensions, not formal proposals. They
are included to demonstrate that the gaps are addressable within the
ISM's existing structure. The use of SHOULD below follows the ISM's
existing convention for conditional controls --- it is illustrative
wording to show how these extensions might read, not normative language.

\textbf{Extension to ISM-0401 (Secure by Design):}

\begin{quote}
\emph{When AI agents are used to generate code for assessed systems, the
organisation's Secure by Design practices SHOULD include
machine-enforceable rules for trust boundary maintenance, defensive
pattern restrictions appropriate to the system's data sensitivity, and
audit trail preservation requirements. Secure by Design principles that
exist only as human-readable documentation are insufficient controls
against AI-generated code, which does not read documentation.}
\end{quote}

\textbf{Extension to ISM-2060/2061 (Code Review and Security-Focused
Peer Review):}

\begin{quote}
\emph{When AI agents generate a significant proportion of code changes,
the organisation SHOULD demonstrate that its code review process
(ISM-2060) and security-focused peer review process (ISM-2061) remain
effective at detecting semantic defects --- not merely syntactic or
conventional defects --- under the volume of changes produced. Evidence
may include measured defect escape rates, review depth audits, or
demonstrated use of automated semantic pre-screening that reduces the
burden on human reviewers.}
\end{quote}

\textbf{New control (Agent Output Trust Boundary):}

\begin{quote}
\emph{Code generated by AI agents SHOULD be treated as external input
requiring validation at the boundary before integration into assessed
systems. The organisation SHOULD define and document the validation
boundary, including what properties are verified (trust boundary
maintenance, audit trail integrity, error handling appropriateness) and
what evidence demonstrates the validation is effective.}
\end{quote}

\textbf{New control (Code Provenance):}

\begin{quote}
\emph{When AI agents are used in the development of assessed systems,
the organisation SHOULD maintain records of which code was generated by
AI agents, which was human-authored, and which was agent-generated then
human-modified. This provenance metadata supports risk assessment,
incident response, and targeted remediation when systematic
agent-introduced defects are discovered.}
\end{quote}

\hypertarget{nist-secure-software-development-framework-ssdf}{%
\subsection{6.2 NIST Secure Software Development Framework
(SSDF)}\label{nist-secure-software-development-framework-ssdf}}

SP 800-218 defines practices for secure software development organised
into four groups. The analysis below maps each group to agentic code
concerns:

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.2830}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.3019}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4151}}@{}}
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Practice Group
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
SSDF Practices
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Agentic Code Coverage
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
\textbf{Prepare the Organization (PO)} & Define security requirements,
roles, training & Does not address training requirements for reviewing
agent output (which requires different skills than reviewing human
output) or organisational capacity planning for agent-scale review
volume \\
\textbf{Protect the Software (PS)} & Protect code, integrity
verification & Addresses integrity of code artefacts but not the trust
classification of code based on its generation method. An
agent-generated commit and a human-authored commit are indistinguishable
in the VCS \\
\textbf{Produce Well-Secured Software (PW)} & Design, code review,
testing & The most relevant group. Practice PW.5 (secure coding
practices), Practice PW.7 (code review), and Practice PW.8 (testing) all
partially apply. However, PW assumes a human developer who can be
trained, who learns from feedback, and whose error rate is independent
across functions --- none of which hold for agents \\
\textbf{Respond to Vulnerabilities (RV)} & Vulnerability response,
disclosure & Does not address the correlated nature of agent-introduced
defects. Standard vulnerability response treats each finding
independently; agent defects require pattern-wide remediation (Section
9.7) \\
\end{longtable}

\textbf{Key SSDF gap:} Practice PW.1 (``Design Software to Meet Security
Requirements and Mitigate Security Risks'') includes task PW.1.1, which
recommends ``using forms of risk-based analysis to determine how much
effort is adequate'' for security practices. This implicitly assumes
that risk is assessable per-component. Agent-generated code introduces
\emph{systematic} risk across many components from a single source ---
the analysis framework needs to account for correlation, not just
per-component risk.

\textbf{NIST's own recognition of this gap:} NIST published SP 800-218A
(July 2024) as a supplement to the SSDF specifically for generative AI
and dual-use foundation model development contexts, acknowledging that
the original framework's human-centred assumptions need AI-specific
augmentation. However, SP 800-218A's focus is on secure practices for AI
\emph{model} development across the SDLC --- not on the assurance of
source code \emph{generated by} AI systems. This is precisely the gap
this paper addresses: substantial guidance now exists for building AI
safely, but almost none for securing what AI builds.

\hypertarget{essential-eight}{%
\subsection{6.3 Essential Eight}\label{essential-eight}}

The Essential Eight maturity model does not directly address software
development practices. However, two strategies are relevant by analogy:

\textbf{Application Control} establishes that not all software should be
trusted equally based on its source. The maturity levels (ML1: prevent
execution of unapproved programs; ML2: restrict to approved directories;
ML3: comprehensive control) provide a model for graduated trust that
could be extended to code generation sources. Analogy: agent-generated
code is ``unapproved software'' until it passes through a validation
boundary --- similar to how an unsigned binary is untrusted until it
meets the application control policy.

\textbf{Restrict Administrative Privileges} establishes the principle of
least privilege. Applied to agentic coding: agents should not have the
ability to modify security-critical configuration (e.g., allowlists,
audit configuration, access control rules) without human approval. This
maps directly to the \texttt{CODEOWNERS} protection and temporal
separation mechanisms described in Appendix B.

\hypertarget{owasp-and-industry-guidance}{%
\subsection{6.4 OWASP and Industry
Guidance}\label{owasp-and-industry-guidance}}

\textbf{OWASP Top 10 for LLM Applications (2025)} primarily addresses
threats \emph{to} LLM systems --- prompt injection, training data
poisoning, model denial of service. The closest entry to this paper's
concerns is \textbf{LLM02 (Insecure Output Handling)}, whose attack
scenarios explicitly include LLM-generated code introducing
vulnerabilities such as SQL injection. However, even LLM02 frames this
as an application-level output-handling problem --- advising developers
to treat LLM output with ``zero-trust'' validation --- rather than
providing a comprehensive treatment of the distinct failure
characteristics of AI-assisted code generation (correlated defects,
review capacity exhaustion, context-inappropriate patterns). The project
has since evolved into the broader \textbf{OWASP GenAI Security
Project}, covering LLM applications, agentic AI systems, and AI-driven
applications --- but no OWASP project specifically targets the assurance
of AI-generated source code in government systems in the sense addressed
by this paper.

\textbf{OWASP Secure Coding Practices} provides a checklist of defensive
coding practices. Ironically, several ``secure'' practices in the OWASP
checklist are precisely the anti-patterns that the agentic threat model
identifies as dangerous in high-assurance contexts. For example,
``validate all input'' is correct, but ``provide a default value when
input is missing'' is context-dependent --- in audit-critical systems, a
missing value should crash, not default. This illustrates the gap
between generic secure coding guidance and domain-specific trust
boundary requirements.

\textbf{MITRE ATT\&CK and CWE} provide taxonomies for attack techniques
and code weaknesses respectively. The agentic code failure modes in
Appendix A do not map cleanly to existing CWE entries because they are
not individual weaknesses --- they are \emph{patterns} that are correct
in most contexts and dangerous in specific ones. A \texttt{.get()} with
a default value is not a weakness; it is a weakness \emph{when applied
to audit-critical data in a system that requires crash-on-corruption}.
Context-dependent weaknesses are not well-served by context-free
taxonomies.

\hypertarget{the-gap-between-securing-ai-and-securing-what-ai-builds}{%
\subsection{6.5 The Gap Between ``Securing AI'' and ``Securing What AI
Builds''}\label{the-gap-between-securing-ai-and-securing-what-ai-builds}}

Across all current frameworks, there is a consistent structural gap:
substantial guidance exists for securing AI systems themselves (the
model, the training pipeline, the inference infrastructure), but almost
no guidance exists for securing systems that AI systems build or modify.

This gap is not surprising --- agentic coding is a recent capability and
guidance takes time to develop. But the gap is widening faster than it
is closing, because:

\begin{itemize}
\tightlist
\item
  Agent adoption is accelerating (driven by generation velocity gains
  and expanded capability --- Section 1.2, Section 8)
\item
  The failure modes are subtle (they pass all existing automated checks
  --- Section 2.2)
\item
  The vocabulary for discussing these failures doesn't exist yet in
  policy contexts (this paper's taxonomy is a first attempt)
\end{itemize}

\hypertarget{detection-coverage-is-worst-where-risk-is-highest}{%
\subsection{6.6 Detection Coverage Is Worst Where Risk Is
Highest}\label{detection-coverage-is-worst-where-risk-is-highest}}

Before listing the structural gaps, it is worth stating the detection
picture plainly. The ACF taxonomy (Appendix A) catalogues thirteen
agentic code failure modes. Of those thirteen:

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4444}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.1944}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.3611}}@{}}
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Detection Level
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Count
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Implication
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
\textbf{None} --- no existing tool detects it & 5 & Requires new tooling
or new review practices \\
\textbf{Partial} --- some tools catch some cases & 5 & Existing tools
provide incomplete coverage \\
\textbf{Good} --- existing tools generally catch it & 1 & Already
addressed by current tooling \\
\textbf{N/A} --- process threat, not code pattern & 2 & Requires process
controls, not technical controls \\
\end{longtable}

Five of thirteen failure modes have zero tool coverage. Both
Critical-rated entries --- ACF-T1 (trust tier conflation) and ACF-E1
(implicit privilege escalation) --- are among the five with no detection
capability whatsoever. The highest-risk failures are precisely the ones
current tooling misses entirely. The full detection breakdown with
specific failure IDs is in Appendix A.

This is the gap in quantitative terms. The structural gaps follow.

\hypertarget{structural-gaps}{%
\subsection{6.7 Structural Gaps}\label{structural-gaps}}

No current framework provides:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  \textbf{A taxonomy of agentic code failure modes} grounded in
  established threat modelling (STRIDE or equivalent)
\item
  \textbf{Controls for semantic correctness} beyond syntactic and
  functional correctness --- trust boundary maintenance, audit trail
  integrity, context-appropriate error handling
\item
  \textbf{Controls for review effectiveness at scale} --- not just ``is
  code reviewed?'' but ``does the review process remain effective at
  agent-generated volume?''
\item
  \textbf{Trust classification for agent output} --- how should
  agent-generated code be treated in the system's trust model?
\item
  \textbf{Accreditation criteria for agentic development workflows} ---
  what evidence must organisations provide to demonstrate that agentic
  coding maintains the required security posture?
\item
  \textbf{Vocabulary for context-dependent code weaknesses} --- patterns
  that are correct in general but dangerous in specific security
  contexts
\item
  \textbf{Correlated failure risk models} --- testing and remediation
  strategies that account for the non-independent failure distribution
  of agent-generated code
\item
  \textbf{Governance perimeter expansion} --- controls for executable
  logic produced by non-developers using agentic tools outside
  traditional SDLC channels (Section 1.2.8)
\end{enumerate}

Current frameworks scope software development controls to recognised
development teams and established code repositories. Agent-generated
automations, integrations, and plugins produced by analysts and
operators outside these channels are not addressed by any current
guidance.

\hypertarget{the-response-landscape}{%
\section{7. The Response Landscape}\label{the-response-landscape}}

The responses available to organisations fall into three categories of
increasing assurance strength (ordered weakest to strongest):

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.1837}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2653}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2041}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.3469}}@{}}
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Control Type
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Mechanism
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Strength
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Example
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
\textbf{Behavioural} & Relies on individual compliance & Weakest ---
requires sustained restraint against incentives & ``Developers should
not run more than one agent concurrently'' \\
\textbf{Procedural} & Relies on organisational process & Moderate ---
requires consistent enforcement and audit & ``Parallel agent-generated
changes require separate review queues and staged approval'' \\
\textbf{Technical} & Constrains the environment & Strongest --- operates
regardless of individual behaviour & ``The CI/CD pipeline enforces
concurrency limits, sequencing rules, or protected-branch gates for
agent-originated changes'' \\
\end{longtable}

Most organisations will implement behavioural controls, aspire to
procedural controls, and underinvest in technical controls --- because
technical controls constrain the velocity that motivated adoption. The
key insight from security engineering applies here: \textbf{controls
that shape the environment are stronger than controls that depend on
restraint.} A rule that developers must not bypass review is an
aspiration; a pipeline that physically prevents unreviewed code from
reaching protected branches is a control.

The sections below are ordered from weakest to strongest assurance, not
from least to most important. All three have a role, but assurance
should not rest primarily on behavioural or procedural controls where
technical enforcement is feasible. Organisations that rely on
behavioural and procedural controls without technical enforcement should
understand that their assurance argument rests on sustained human
compliance with rules that run directly against the productivity
incentive that makes agentic development attractive.

\hypertarget{process-controls-strengthening-existing-practices}{%
\subsection{7.1 Process Controls (Strengthening Existing
Practices)}\label{process-controls-strengthening-existing-practices}}

\textbf{Current best practice, adapted for agentic velocity:}

\hypertarget{enhanced-code-review-protocols}{%
\subsubsection{Enhanced code review
protocols}\label{enhanced-code-review-protocols}}

Mandate security-focused review (not just correctness review) for
agent-generated code. Require reviewers to attest that trust boundaries
were verified, not just that the code ``looks right.'' This is a process
change, not a technology change, but it requires explicit recognition
that agent code needs different review criteria than human code.

Specifically, review checklists for agent-generated code should include:

\begin{itemize}
\tightlist
\item
  Were trust boundaries maintained? (Does external data pass through
  validation before internal use?)
\item
  Are error handlers audit-preserving? (Do \texttt{except} blocks
  propagate to the audit system, or swallow and continue?)
\item
  Are default values justified? (Does every \texttt{.get()} or
  \texttt{getattr} with a default represent a legitimate design
  decision, or a fabrication of missing data?)
\item
  Is the code's failure mode correct for the context? (Should this code
  crash, quarantine, or continue on error?)
\end{itemize}

\hypertarget{separation-of-generation-and-review}{%
\subsubsection{Separation of generation and
review}\label{separation-of-generation-and-review}}

The person (or agent) who generates the code must not be the sole
reviewer. This already applies to human-authored code in most government
contexts; extending it to agent-generated code means ensuring that agent
self-review (e.g., an agent checking its own output) does not count as
an independent review. This has a subtlety: multi-agent workflows where
one agent generates code and another reviews it using different models
are not independent review in the statistical sense --- the models share
overlapping training corpora and failure modes. However, a more nuanced
form of multi-agent review offers meaningful value.

\hypertarget{structured-perspective-diversity-in-agent-assisted-review}{%
\subsubsection{Structured perspective diversity in agent-assisted
review}\label{structured-perspective-diversity-in-agent-assisted-review}}

While model diversity (using different models) provides limited
independence, \emph{perspective diversity} --- prompting the same model
with different analytical frames --- can produce meaningfully different
coverage. One useful way to organise these perspectives is by
\emph{cognitive function} rather than by domain. Illustrative functions
include: an \textbf{architectural reviewer} (``Is the shape right?''), a
\textbf{problem-framing reviewer} (``Is this solving the right problem
at the right location?''), an \textbf{implementation reviewer} (``Was it
implemented correctly?''), and a \textbf{quality reviewer} (``Are the
tests and verification strategy adequate?''). These lenses are not
independent, but they surface different classes of issue. A trust
boundary violation (ACF-T1) may be an implementation defect; the
\emph{reason} it exists may be an architectural misplacement --- a
validation that belongs in a different layer; and the tests may still
pass while verifying the wrong thing. The problem-framing perspective is
particularly underexplored: evaluating whether a change addresses the
root cause or patches a symptom often requires holding broader
dependency context in working memory --- something agents may assist
with and time-pressured human review often struggles to sustain.

This is not independence --- the underlying model's blind spots persist
across all frames --- but it is \emph{faceted analysis} that surfaces
different failure classes. In practice, a small set of prompted
perspectives may provide broader first-pass coverage at agent speed than
a single undifferentiated review pass, reducing the volume of issues
that reach human review while increasing the diversity of issues caught.
Organisations should develop \emph{role-specific review prompts} aligned
to their threat model and architecture --- not generic ``review this
code'' instructions. The prompts themselves become reviewable,
version-controlled security artefacts: institutional knowledge encoded
in a different form than the machine-readable rules described in Section
7.2, but serving a complementary function.

The honest caveat: this is still a procedural control, not a technical
one. It depends on someone actually running the prompted reviews, and
the quality depends on prompt design. It sits in the middle tier of the
control strength hierarchy in this section. But it is achievable now
with no tooling investment, which makes it a useful bridge while
automated enforcement tooling matures.

\hypertarget{volume-aware-review-capacity-planning}{%
\subsubsection{Volume-aware review capacity
planning}\label{volume-aware-review-capacity-planning}}

If agents materially increase code generation volume, review capacity
must be addressed --- either through additional reviewers, automated
pre-screening that reduces the human review burden, or rate-limiting
agent output to match review capacity. Ignoring the volume mismatch
means the review control degrades silently.

\hypertarget{project-level-instructions-as-a-generation-time-control}{%
\subsubsection{Project-level instructions as a generation-time
control}\label{project-level-instructions-as-a-generation-time-control}}

Most agent frameworks support project-level configuration --- system
prompts, instruction files, memory stores --- that shape agent behaviour
within a session. These instructions can encode project-specific rules
(``never use \texttt{.get()} on audit data,'' ``all error handlers for
audit-write operations must propagate, not catch-and-continue'') and
reduce the frequency of the failure modes in Appendix A at generation
time. This is a behavioural control (weakest tier): it depends on the
agent respecting the instructions, the instructions being comprehensive
enough to cover all failure modes, and the agent not generalising
incorrectly from specific rules. Instructions that say ``don't use
\texttt{.get()} on audit data'' do not teach the agent the underlying
principle ``don't fabricate defaults where data absence is meaningful''
--- every specific rule must be spelled out. Nonetheless, project-level
instructions are an immediately deployable, zero-cost control that
measurably reduces (without eliminating) the volume of semantic
violations reaching the review pipeline.

\hypertarget{provenance-tracking-for-agent-output}{%
\subsubsection{Provenance tracking for agent
output}\label{provenance-tracking-for-agent-output}}

Organisations should maintain records of which code was generated by
agents, which was human-authored, and which was agent-generated then
human-modified. This metadata is relevant for both security assessment
(understanding the trust profile of different code regions) and for
incident response (when a defect is found, knowing whether it originated
from agent generation helps diagnose the failure mode).

\hypertarget{technical-controls-whats-buildable}{%
\subsection{7.2 Technical Controls (What's
Buildable)}\label{technical-controls-whats-buildable}}

When organisations evaluate the feasibility of security tooling for
agentic code, the instinctive question is: \emph{``How big is it? How
many lines of code? How many hours to build?''} This is the wrong
question. It inherits the assumption that scope correlates with
assurance --- that a 200-line tool provides less assurance than a
2000-line tool, or that 220 hours of development is insufficient for a
security-critical control.

The right question is: \textbf{``How do you know it's correct?''}

This question applies recursively. If the security enforcement tool is
itself built by an agent (which is increasingly likely for any new
tool), then the tool's correctness is subject to the same threat model
it exists to address. The verification mechanism --- not the
implementation size --- is the assurance argument.

This is not a new problem. Test suites that consolidate object
construction into shared factories face the same trade-off: centralising
institutional knowledge (what a valid test object looks like) eliminates
a class of bugs caused by ad hoc misconfiguration, but concentrates risk
in the factories themselves. A defect in a factory propagates to every
test that depends on it. The engineering response is not to avoid
centralisation --- the alternative is worse --- but to verify the
central artefacts with disproportionate rigour. The same principle
applies to any tool that encodes institutional security knowledge: the
value of centralisation is high, but so is the cost of getting the
central artefact wrong.

This reframing has direct implications for policy:

\begin{itemize}
\tightlist
\item
  \textbf{Accreditation should evaluate the verification story}, not the
  implementation scope. A small tool with a rigorous golden corpus,
  self-hosting gate, and measured precision is stronger than a large
  tool without these properties.
\item
  \textbf{``How do you know the agent wrote it correctly?''} is the
  question that applies to all agent-generated code, including the tools
  that check agent-generated code. The answer must be grounded in
  independent verification (test corpus, self-hosting, measured false
  positive/negative rates), not in the development process used to
  produce it.
\item
  \textbf{The line between ``tool'' and ``assurance argument'' blurs.}
  The design of a semantic boundary enforcer (Appendix B) is essentially
  a structured answer to ``how do you know the code is correct?'' ---
  and that answer is valuable independent of whether the specific tool
  gets built.
\end{itemize}

This reframing directly informs Recommendations 7 and 8 ---
accreditation should evaluate the verification story, and IRAP assessors
should assess review quality, not just review process. The technical
controls below should be evaluated on this basis.

\textbf{Automated semantic boundary enforcement.} Static analysis tools
that verify data provenance and trust tier flow --- not just type shape
--- at the code level. These tools check that external data passes
through validation boundaries before reaching internal processing, that
error handling preserves audit trails, and that defensive patterns are
not used on data that should crash on anomaly.

\textbf{Technical feasibility finding:} The case study codebase (Section
8) runs a pattern-matching enforcement gate that catches trust boundary
violations in CI. Building on this production experience, an enhanced
semantic boundary enforcer for Python has been designed (see Appendix B)
with the following properties:

\begin{itemize}
\tightlist
\item
  Zero external dependencies (uses only Python's standard library AST
  module)
\item
  Works with standard Python --- no language modifications, no runtime
  dependencies
\item
  Produces findings in SARIF format (industry standard for static
  analysis results)
\item
  Can operate as a pre-commit check, CI gate, or agent self-check before
  code submission
\item
  Verification properties (golden corpus, self-hosting gate, measured
  precision) are independently auditable --- see Appendix B, Section B.5
\end{itemize}

This is not the only possible technical control, but it demonstrates
that the problem is tractable. Existing static analysis platforms ---
Semgrep (custom rule authoring), CodeQL (dataflow analysis), and
Pysa/Pyre (taint tracking) --- could be extended with rules targeting
the ACF taxonomy's failure modes, particularly trust tier flow (ACF-T1)
and validation boundary gaps (ACF-T2). The design in Appendix B chose
AST-level analysis with zero dependencies for deployment simplicity, but
organisations with existing static analysis infrastructure should
evaluate whether their current tools can be adapted before building new
ones.

\textbf{Agent-assisted semantic analysis.} The prompted perspective
diversity described in Section 7.1 applies not only to reviewing changes
but to \emph{analysing existing code}. An agent prompted with a specific
analytical frame can perform a cold read of a source file --- tracing
execution flow, following data through trust boundaries, and evaluating
error-handling paths --- with a level of context-sensitive analysis that
traditional static analysis often cannot match and that human reviewers
cannot sustain at scale. Static analysers are strongest at structural,
syntactic, and formally encoded dataflow patterns; they cannot reliably
determine whether a \texttt{.get()} default is \emph{contextually
appropriate} without project-specific semantic rules. Human reviewers
can make that judgement, but cannot practically sustain file-by-file
semantic tracing across multiple analytical lenses at codebase scale.
Agent-assisted analysis therefore occupies a previously sparse middle
ground: semantic inspection applied at a scale that was economically
impractical for human review alone. This is not a replacement for either
static analysis or human review --- it catches a different class of
issue from both, and its blind spots differ from both. In most settings,
it is likely to be most useful as a non-blocking discovery control,
complementing CI-integrated static analysis and targeted human review
rather than replacing either.

In the case study deployment (Section 8), periodic full-codebase agent
crawls --- every file analysed through prompted analytical frames ---
routinely surfaced dozens of findings per pass, the majority of which
would not have been caught by conventional static analysis, incremental
code review, or unprompted agent review. The findings emerge
specifically \emph{because} the analytical frames are non-default ---
neither human reviewers under time pressure nor agents given generic
review instructions naturally adopt the perspective of, say, an
architectural reviewer evaluating layer responsibility or a
problem-framing reviewer asking whether a fix addresses root cause or
symptom. This points to a human limitation that is distinct from time
pressure: \emph{cognitive range}. A senior backend engineer cannot
practically adopt a PyTorch engineer's analytical frame, or a security
architect's, or a data quality auditor's --- not because they lack
intelligence, but because genuine expertise in each frame takes years to
develop. Prompted agents do not have deeper expertise in any single
frame than a domain specialist, but they can adopt multiple frames
without the switching cost and domain-knowledge barriers that make it
impractical for any individual human reviewer to cover more than one or
two perspectives well. In this sense, prompted agents are
\emph{polymorphic} reviewers --- not deeper than a specialist in any
single frame, but able to provide breadth of analytical coverage that no
individual human can match regardless of time budget. Most findings are
individually low-impact --- edge cases in cold paths, minor deviations
from stated invariants, inconsistencies that do not currently trigger in
testing --- but they represent legitimate deviations from the codebase's
architectural rules, and in an auditable system, each deviation is a gap
in the assurance argument regardless of its current exploitability. The
cost is non-trivial (such crawls consume substantial API quota), but the
defect yield suggests the economics may favour periodic comprehensive
analysis over exclusive reliance on per-change review.

\textbf{Key architectural principle --- ``parasitic, not parallel'':}
Effective tools for this space must extend existing programming language
machinery (annotations, type hints, decorators) rather than creating
parallel systems that require adoption of new syntax or tools. Tools
that require developers to learn a new language or adopt a new framework
face adoption resistance that undermines their security value.
Critically, enforcement must live inside the existing CI/CD pipeline ---
pre-commit hooks, CI gates, pull request checks --- not in a separate
workflow. If a security tool slows down the very velocity the
organisation bought the AI agent to achieve, the tool will be bypassed.
The enforcement mechanism succeeds by being invisible to the fast path
and blocking only on genuine violations.

\hypertarget{policy-controls-what-doesnt-exist-yet}{%
\subsection{7.3 Policy Controls (What Doesn't Exist
Yet)}\label{policy-controls-what-doesnt-exist-yet}}

\textbf{Standardised vocabulary.} The taxonomy in this paper (Section 3
and Appendix A) provides a starting point. Government cybersecurity
guidance needs terminology for agentic code failure modes ---
``competence spoofing,'' ``trust tier conflation,'' ``audit trail
destruction through defensive patterns'' --- that practitioners can use
in security assessments, risk registers, and accreditation
documentation.

\textbf{Accreditation criteria for agentic development workflows.} IRAP
assessments and similar accreditation processes need criteria for
evaluating whether an organisation's use of AI coding agents maintains
the security posture required by the system's classification. This
includes:

\begin{itemize}
\tightlist
\item
  How agent output is validated before integration
\item
  How review effectiveness is maintained under volume pressure
\item
  How trust boundaries are verified in agent-generated code
\item
  What attestation is required from human reviewers
\end{itemize}

\textbf{Agent output classification.} A formal determination of how
agent-generated code should be treated in the trust model --- as
external input (Tier 3), as tool output (requiring specific validation),
or as a new category requiring its own controls.

\textbf{Extend SDLC-equivalent controls to executable logic produced
outside formal development teams.} Organisations should inventory and
govern agent-generated plugins, automations, BI extensions, workflow
scripts, low-code components, and similar artefacts produced by
analysts, operators, and other non-developer staff (see Section 1.2.8).
Where such artefacts affect trust boundaries, access control, audit
trails, or data integrity, they should be subject to provenance, review,
and validation controls proportionate to their impact --- even when the
producers do not consider themselves developers and the artefacts do not
live in a formal version control system. This is the policy response to
the governance perimeter problem: the SDLC boundary has expanded, and
controls must follow it.

\hypertarget{case-study-agentic-development-under-compliance-constraints}{%
\section{8. Case Study: Agentic Development Under Compliance
Constraints}\label{case-study-agentic-development-under-compliance-constraints}}

\emph{This section presents a composite, de-identified account drawn
from experience with agentic development in compliance-constrained
environments. Specific implementation details have been generalised. The
system and tooling described are open-source, Commonwealth-owned assets;
they are de-identified here to keep focus on the generalisable threat
model rather than implementation-specific design choices.}

\hypertarget{context}{%
\subsection{8.1 Context}\label{context}}

An engineering team developing an auditable data processing platform ---
a system where every decision must be traceable to its source data,
configuration, and code version. The system processes sensitive data
under compliance requirements that mandate complete audit trails, data
integrity verification, and defence-in-depth security controls.

The team adopted AI coding agents as a primary development tool, with
agents generating the majority of new code. The codebase enforces strict
architectural rules: a tiered trust model for data handling, mandatory
crash-on-corruption for internal data, quarantine-and-continue for
external data, and no defensive programming patterns (no \texttt{.get()}
on typed objects, no bare \texttt{except}, no silent error swallowing).

These rules are documented extensively but are \textbf{institutional
knowledge} --- they exist in project documentation, not in the
programming language. Python permits all of the patterns the project
forbids.

\hypertarget{the-compliance-tax}{%
\subsection{8.2 The Compliance Tax}\label{the-compliance-tax}}

Compliance requirements impose substantial overhead on the development
workflow:

\textbf{Hash integrity verification.} Every data operation produces
cryptographic hashes for audit trail integrity. Changes to hashing
logic, serialisation formats, or data structures require recompilation
and verification of hash chains. This is mechanical work that agents
handle well, but the \emph{verification} that the agent produced correct
hashes requires human review --- or automated checking.

\textbf{Audit trail completeness.} Every code path must produce audit
records. Missing audit records are treated as evidence tampering, not as
bugs. This means error handlers cannot swallow exceptions, partial
operations must roll back, and ``log and continue'' is not an acceptable
failure mode. Agents consistently produce error handlers that violate
these requirements because their training data overwhelmingly models
``log and continue'' as best practice (Section 2.5).

\textbf{Trust boundary enforcement.} The team maintains a static
analysis tool (manually built, project-specific) that detects defensive
coding patterns in contexts where crash-on-corruption is required. This
tool has an allowlist mechanism for legitimate exceptions. Maintaining
the allowlist, reviewing new findings, and ensuring agents don't
introduce violations that the tool doesn't yet detect is a recurring
cost.

\textbf{The overhead is real --- the team estimates the aggregate
compliance overhead at 15-25\% of total development time} --- but its
distribution is uneven and reveals something about how agents experience
unfamiliar workflows. On large changes (new features, multi-file
refactors), compliance overhead is trivially small relative to the work
--- even if the agent spends 10 minutes recalculating 500 allowlist
hashes after a major refactor, that is a reasonable investment against
hours of productive generation. But on small changes, the ratio inverts
dramatically. An agent fixing a one-line bug spends 30 seconds on the
fix and 60 seconds grappling with the CI pipeline: discovering that an
enforcement gate is blocking the commit, parsing the error to understand
what the tool expects, calculating a new allowlist hash, and editing the
configuration. The compliance work takes twice as long as the actual
work. The agent still completes the entire cycle faster than a human
would, but the unfamiliar workflow --- one the agent has never seen in
training data and must rediscover each session --- imposes an outsized
cost on small tasks, and this skew toward small-change cases is where
the bulk of the overhead concentrates.

This is not new overhead introduced by agentic coding. It is the
\emph{same compliance overhead} that existed before agents,
redistributed. Before agents, humans spent that time writing compliant
code slowly. With agents, humans spend it reviewing agent code for
compliance quickly. The total compliance cost is similar; the
development velocity is dramatically higher.

\hypertarget{productivity-gains-despite-friction}{%
\subsection{8.3 Productivity Gains Despite
Friction}\label{productivity-gains-despite-friction}}

Despite the compliance overhead, the team reports substantial
productivity gains from agentic coding:

\begin{itemize}
\tightlist
\item
  \textbf{Mechanical refactoring} (renaming, restructuring, pattern
  application across files) is handled almost entirely by agents. This
  work is tedious for humans but trivial for agents, and the compliance
  requirements don't make it harder --- the agent produces the
  refactoring, the existing test suite and CI pipeline verify it.
\item
  \textbf{Boilerplate generation} (new plugins, test scaffolding,
  configuration structures) is dramatically accelerated. The agent
  follows the project's existing patterns, and the structural
  correctness is verifiable by tests.
\item
  \textbf{Bug investigation and test writing} benefit from agents'
  ability to rapidly explore code paths and generate test cases. The
  agent's lack of institutional knowledge is less of a liability when
  the task is ``write a test that exercises this code path'' rather than
  ``write code that maintains trust boundaries.''
\end{itemize}

The pattern is clear: \textbf{agents excel at tasks where correctness is
structurally verifiable} (tests pass, types check, linter is clean) and
struggle at tasks where \textbf{correctness requires institutional
knowledge} (trust boundary maintenance, audit trail completeness,
appropriate error handling in compliance contexts).

\hypertarget{where-the-current-process-fails}{%
\subsection{8.4 Where the Current Process
Fails}\label{where-the-current-process-fails}}

The team has observed agent-generated code of questionable quality
passing human review. In each case, the code was:

\begin{itemize}
\tightlist
\item
  Syntactically correct
\item
  Consistent with project conventions
\item
  Passing all existing tests
\item
  Semantically wrong in the compliance context
\end{itemize}

The failure modes map directly to the taxonomy in Section 3:

\textbf{Competence spoofing (ACF-S1).} Agent generates \texttt{.get()}
with a default value on a data structure where a missing field indicates
a critical failure in an upstream internal component --- absence is
evidence of corruption, not a case to handle gracefully. The code
appears defensive and robust. A reviewer under time pressure sees
``handles the missing case'' rather than ``fabricates data where absence
means something has gone seriously wrong upstream and continuing with a
default is irresponsible.''

\textbf{Audit trail destruction (ACF-R1).} Agent wraps an audit-critical
operation in a \texttt{try/except} that logs the error and continues.
The code appears to handle errors gracefully. The reviewer doesn't
recognise that the caught exception should propagate to the audit system
rather than being logged and swallowed.

\textbf{Trust tier conflation (ACF-T1).} Agent deserialises data from an
external API and passes it directly to an internal processing function.
The code appears clean --- no obvious security issues. The reviewer
doesn't see the missing validation boundary because both the external
data and internal data are the same Python type (\texttt{dict}).

In each case, the defect was caught later --- by the project's existing
static analysis tool, by a different reviewer examining adjacent code,
or by a test failure in a downstream component that received malformed
data. The initial review process, which was supposed to catch these
issues, had signed off.

The project's automated semantic boundary enforcement tool (Section 7.2;
see Appendix B for the proposed enhanced design) provides a rough
empirical signal. In steady-state development on a
\textasciitilde80,000-line Python codebase with agents generating the
majority of new code, the CI gate catches approximately 1--2 trust
boundary violations per day --- patterns from the ACF taxonomy
(primarily ACF-S1, ACF-R1, and ACF-T1) that the generating agent
introduced. These violations never enter the repository. They are caught
at the pre-commit or CI boundary before a human reviewer ever sees them.
Without the gate, they would enter the codebase through normal code
review, because they look like correct defensive Python --- which is
exactly what makes them dangerous (Section 2.3). Over a quarter, this
accumulates to roughly 90--180 violations that automated tooling caught
at the boundary.

Three aspects of this number merit attention. First, it represents only
the violations the tool detects --- the tool's coverage of the ACF
taxonomy is incomplete (Appendix A, Detection Capability Summary), so
the true rate of semantic violations is higher. Second, and more
important for the policy argument: these are not exotic edge cases. They
are the routine, daily output of a well-prompted agent operating on a
well-documented codebase with project-level instructions that explicitly
prohibit the patterns in question. The agent still produces them,
because the patterns are deeply embedded in its training data --- and
because the agent has no persistent memory across sessions (Section
2.4), the same patterns recur regardless of how many times they have
been caught previously. Third, the enforcement model is not advisory ---
it is a gate with an allowlist. A pattern flagged by the enforcer either
gets fixed by the agent or requires a human-authored exception with a
rationale, a reviewer identity, and an expiry date (Appendix B, Section
B.4). Legitimate uses of otherwise-restricted patterns go through;
unconscious pattern completion from training data does not.

The question for organisations without this kind of enforcement is not
whether these patterns exist in their agent-generated code --- it is
whether anything is catching them.

\hypertarget{the-redirection-insight}{%
\subsection{8.5 The Redirection Insight}\label{the-redirection-insight}}

The team's experience reveals that automated semantic enforcement
doesn't \emph{add} tedium --- it \textbf{redirects existing tedium}
toward higher-value activities.

Without automated enforcement, humans manually review every agent output
for trust boundary violations. This is:

\begin{itemize}
\tightlist
\item
  \textbf{Error-prone:} The failure modes look like correct code
  (Section 3)
\item
  \textbf{Fatigue-inducing:} Reviewing dozens of agent-generated
  functions per day for subtle semantic violations degrades review
  quality (Section 4.2)
\item
  \textbf{Unscalable:} As agent velocity increases, review capacity
  doesn't
\end{itemize}

With automated enforcement, the machine catches structural trust
boundary violations (unterminated taint paths, defensive patterns on
typed objects, missing validation boundaries). Humans focus on
\textbf{semantic issues that require institutional knowledge} ---
whether the trust topology is correctly declared, whether the validation
logic is actually correct (not just structurally present), whether the
audit trail captures the right information.

This is a genuine improvement in security posture, not just efficiency:

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.2653}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.3878}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.3469}}@{}}
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Review Focus
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Without Automation
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
With Automation
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
``Is \texttt{.get()} used on typed objects?'' & Human scans for pattern
(error-prone) & Machine catches structurally (reliable) \\
``Does this error handler preserve the audit trail?'' & Human evaluates
(moderate difficulty) & Machine flags broad \texttt{except} blocks;
human evaluates the specific cases \\
``Is the trust topology correctly declared for this new module?'' &
Human evaluates (requires institutional knowledge) & Human evaluates (no
change --- this is irreducibly human) \\
``Is this validation function actually validating?'' & Human evaluates
(requires domain knowledge) & Machine checks structural presence of
control flow; human evaluates semantic adequacy \\
\end{longtable}

The total review burden may be similar, but the \textbf{distribution of
human attention shifts} from low-value pattern scanning to high-value
semantic evaluation. The compliance tax is the same; the assurance yield
is higher.

A less obvious but equally important effect: \textbf{more compliance
work is actually executed in an agentic workflow than in a purely human
one.} The same property that makes agents dangerous --- no persistent
learning, no internalised shortcuts --- makes them unusually compliant
enforcement subjects. A developer fixing a one-line bug at 4:55pm on a
Friday is not going to recalculate 500 allowlist hashes, rerun the full
CI pipeline, and write a whitelist rationale for the \texttt{.get()}
they just introduced. They will commit, push, and deal with it Monday.
The agent doesn't have a Friday. It doesn't have a deadline. It lacks
the judgment to decide ``this governance step isn't worth it for a
one-liner.'' It just does the work. Every session it hits the pre-commit
hook fresh, grapples with it fresh, and either satisfies it or fails. It
never develops the human instinct of ``I'll fix that later'' --- because
there is no later for it.

This is the mirror image of the persistent-learning problem described in
Section 2.4. The agent doesn't learn your security rules, but it also
doesn't learn which rules it can get away with skipping. The net effect
is that the agent pays the governance tax that humans under deadline
pressure quietly defer. For anyone who has audited a development team
and found the gap between ``documented process'' and ``what actually
happens under delivery pressure,'' this is a significant finding: agents
are simultaneously the worst authors and the best compliance subjects.

This also adds nuance to the control-strength hierarchy in Section 7.
The paper argues behavioural controls are weakest because they require
sustained restraint against incentives. That is true for humans. For
agents, behavioural controls are weak for a \emph{different} reason ---
not because the agent will choose to skip them, but because it won't
remember them next session. Technical controls (CI gates, pre-commit
hooks) are strong for agents for the same reason they are strong for
humans --- they are environmental, not volitional --- but they gain an
additional benefit: the agent won't resent the gate, won't lobby to have
it removed, won't find a workaround, and won't develop learned
helplessness about false positives. It will run into the wall, fix the
issue, and move on. Every single time.

The underlying lesson from the case study is this: \textbf{agentic
development is viable precisely because the agent will execute
governance that humans under pressure quietly defer --- but it requires
governance designed for the agent's actual failure modes, not the
human's.} Human governance assumes the author remembers last session's
feedback, can be trained over time, and will exercise judgment about
which rules matter. Agent governance must assume none of those things.
It must be environmental (CI gates, not documentation),
boundary-enforced (pre-commit, not post-review), and stateless (every
session is the first session). Organisations that apply human-shaped
governance to agents will get the agent's compliance without catching
the agent's mistakes. Organisations that design agent-shaped governance
will get both.

\hypertarget{open-questions-for-the-community}{%
\section{9. Open Questions for the
Community}\label{open-questions-for-the-community}}

This paper does not attempt to answer these questions definitively. They
are posed for discussion and community input.

\hypertarget{trust-classification}{%
\subsection{9.1 Trust Classification}\label{trust-classification}}

\textbf{Should agent-generated code be treated as a distinct trust tier
in security assessments?}

The Tier 3 (external/untrusted) classification proposed in Section 5 is
a starting point, but it may be too coarse. Agent code that has been
reviewed and modified by a human developer exists in an intermediate
state --- it originated externally but has been validated. Does this
validation elevate it to Tier 2? What constitutes sufficient validation?

A related question: does the trust classification of agent code change
over time? Code generated by an agent 12 months ago that has been in
production, passed multiple review cycles, and survived real-world use
arguably has a different trust profile than code generated 5 minutes
ago. Should the trust tier be static (based on origin) or dynamic (based
on accumulated evidence)?

\hypertarget{accreditation-burden}{%
\subsection{9.2 Accreditation Burden}\label{accreditation-burden}}

\textbf{What is the appropriate accreditation burden for organisations
using agentic coding in PROTECTED or higher systems?}

Current IRAP assessment criteria do not specifically address AI code
generation. Should they? If so, what evidence should an organisation
provide to demonstrate that its use of agents maintains the required
security posture?

Candidate evidence requirements (for discussion):

\begin{itemize}
\tightlist
\item
  Demonstration that agent output passes through a defined validation
  boundary (Section 5.3)
\item
  Measured review effectiveness metrics (not just ``we have a review
  process'' but ``here is evidence the process catches the failure modes
  in Appendix A'')
\item
  Provenance records showing which code was agent-generated (Section
  7.1)
\item
  Evidence that automated semantic enforcement tools are in use and
  their precision is measured (Appendix B, Section B.5)
\end{itemize}

Must the validation boundary itself be assessed? If an organisation
builds or adopts automated enforcement tooling, does that tooling become
part of the assessed system?

\hypertarget{static-vs.-runtime-assurance}{%
\subsection{9.3 Static vs.~Runtime
Assurance}\label{static-vs.-runtime-assurance}}

\textbf{Can static analysis tools provide sufficient assurance for
agent-generated code, or is runtime verification necessary?}

Static analysis can detect structural patterns (Section 7.2) but cannot
verify semantic correctness in all cases. Runtime verification
(monitoring trust tier flow at execution time) provides stronger
guarantees but introduces performance overhead and operational
complexity. Where is the appropriate trade-off for government systems?

The design work in Appendix B suggests a middle ground: static analysis
as the primary enforcement mechanism with narrow runtime verification in
test mode only --- calibrating the static analyzer's coverage without
introducing production runtime dependencies. Whether this trade-off is
acceptable for a given classification level is a policy question, not a
technical one.

\hypertarget{review-quality-measurement}{%
\subsection{9.4 Review Quality
Measurement}\label{review-quality-measurement}}

\textbf{How do you measure the effectiveness of human review when the
volume of agent output makes comprehensive review impractical?}

Current accreditation processes assume that code review provides a
certain level of assurance. If review quality degrades under volume
pressure (Section 4), the assurance claim is no longer valid. How should
organisations measure and demonstrate that their review process remains
effective?

Possible metrics (all have limitations):

\begin{itemize}
\tightlist
\item
  \textbf{Defect escape rate:} How many agent-introduced defects are
  found post-review? Requires knowing the total defect count, which
  requires a detection mechanism independent of the review process.
\item
  \textbf{Review depth sampling:} Periodically audit review decisions
  for thoroughness. Resource-intensive but directly measures quality.
\item
  \textbf{Automated pre-screening coverage:} What percentage of the
  failure modes in Appendix A are caught by automated tools before
  reaching human review? This doesn't measure review quality directly
  but measures how much the review process is being supplemented.
\item
  \textbf{Time-per-change metrics:} Review duration per lines changed. A
  leading indicator --- if review time per change is declining while
  change volume increases, review quality is likely degrading.
\end{itemize}

\hypertarget{agent-self-regulation}{%
\subsection{9.5 Agent Self-Regulation}\label{agent-self-regulation}}

\textbf{Should agents be required to self-check their output against
security rules before submitting it?}

Many agent workflows already include self-checking loops --- agents run
linters, type checkers, and test suites against their own output before
submission, often triggered by pre-commit hooks or CI integration. This
is current practice, not a future possibility. The question is whether
this practice should be \emph{required} and whether it should extend to
semantic security checks.

Pre-generation self-checking (where the agent validates its own output
against security rules before submission) closes the feedback loop at
the point of maximum leverage --- before the code enters the repository.
This is technically feasible (Appendix B describes a \texttt{-\/-stdin}
mode for exactly this purpose).

But current practice and the proposal both raise questions:

\begin{itemize}
\tightlist
\item
  Does self-checking constitute validation, or must validation be
  independent? The agent checking its own output against structural
  rules is different from the agent evaluating whether its output is
  semantically correct --- the first is a mechanical verification, the
  second is the same judgment that produced the code.
\item
  Can an agent meaningfully check for the failure modes it is
  predisposed to produce? If the agent's training data biases it toward
  \texttt{.get()} with defaults, will it also generate code that
  satisfies a rule against that pattern, or will it generate code that
  evades the rule while preserving the same semantic flaw?
\item
  If self-checking catches 80\% of structural violations before human
  review, does this improve or degrade human review quality? It might
  improve it (reviewers focus on harder problems) or degrade it
  (reviewers assume the pre-check caught everything and reduce
  scrutiny).
\end{itemize}

The answer becomes more nuanced when considering structured perspective
diversity (Section 7.1). A single agent checking its own output is not
meaningful validation --- the same biases that produced the code will
evaluate it. But a \emph{structured ensemble of prompted perspectives}
checking the generating agent's output is a different proposition: the
reviewing agents share the generator's underlying model biases, but
their prompted analytical frames constrain \emph{what they attend to},
producing coverage across different subsets of the failure taxonomy
(Appendix A). This is not independence, but it is orthogonality --- the
blind spots of a security-focused reviewer prompt and a
data-quality-focused reviewer prompt overlap less than two
identically-prompted agents' blind spots would. Whether this constitutes
``validation'' in a formal assurance sense is an open question, but it
is a meaningfully stronger control than single-agent self-review.

\hypertarget{cross-organisational-standards}{%
\subsection{9.6 Cross-Organisational
Standards}\label{cross-organisational-standards}}

\textbf{Should there be a common agentic code security standard across
Australian Government?}

Individual organisations developing their own agentic code policies will
produce inconsistent and potentially conflicting approaches. A common
standard --- even a lightweight one --- would provide:

\begin{itemize}
\tightlist
\item
  A shared vocabulary for discussing agentic code risks (the taxonomy in
  Appendix A is a candidate starting point)
\item
  A minimum bar for controls that all agencies using agentic coding must
  implement
\item
  A basis for mutual recognition of agentic development practices across
  agencies
\item
  Consistency in IRAP assessment criteria for agentic workflows
\end{itemize}

A transparency note is warranted here. The case study project described
in Section 8 catches approximately 1--2 trust boundary violations per
day at its CI gate. It does not currently share those patterns with
other organisations. This paper recommends cross-organisational sharing
of agent-introduced defect patterns, yet the project that generated the
recommendation does not practise it. The reason is prosaic: no mechanism
exists to make participation easy. There is no shared taxonomy for
classifying these findings, no intake channel for reporting them, and no
expectation that other organisations are looking for them. The gap this
section identifies is not hypothetical --- it is precisely the gap this
paper's own practice falls into.

The counterargument: standardisation too early may lock in controls that
prove inappropriate as the technology evolves rapidly. A vocabulary
standard and minimum control set may be more durable than detailed
prescriptive requirements.

\hypertarget{the-correlated-failure-problem}{%
\subsection{9.7 The Correlated Failure
Problem}\label{the-correlated-failure-problem}}

\textbf{How should risk models account for correlated failures in
agent-generated code?}

Traditional software risk models assume that defects are approximately
independent --- a bug in one function doesn't predict a bug in another.
Agent-generated code violates this assumption (Section 2.4). A single
training data bias produces the same failure mode across every function
the agent generates.

This has implications for:

\begin{itemize}
\tightlist
\item
  \textbf{Testing strategy:} Independent sampling (testing a random
  subset of functions) underestimates defect rates when failures are
  correlated. If you find a trust boundary violation in one
  agent-generated function, the probability that the same violation
  exists in other agent-generated functions is much higher than if a
  human had written them.
\item
  \textbf{Risk assessment:} The risk of a single agent-generated defect
  may be low, but the risk of a \emph{systematic} defect affecting
  dozens or hundreds of functions is qualitatively different. How should
  risk registers capture correlated agent failure risk?
\item
  \textbf{Remediation scope:} When a defect pattern is found in agent
  code, remediation should not be limited to the specific instance. The
  entire codebase should be scanned for the same pattern --- because
  correlated failures mean the pattern is likely repeated.
\item
  \textbf{Triage model:} Correlated failures mean 50 firings of the same
  rule across a codebase is one systematic issue requiring a systematic
  fix, not 50 independent tickets. Organisations that triage
  agent-generated defects as independent findings will overwhelm their
  remediation capacity on what is, operationally, a single root cause.
\end{itemize}

\hypertarget{contracted-out-development}{%
\subsection{9.8 Contracted-Out
Development}\label{contracted-out-development}}

\textbf{How does the threat model apply when most code is written by
contractors, not agencies?}

The majority of software development for Australian Government systems
is performed by contracted service providers, not by agency staff. This
paper's threat model applies to contractors' agents with the same force
--- an agent operated by Accenture, Leidos, or a boutique consultancy
produces the same correlated, context-inappropriate patterns as an agent
operated in-house. But the governance implications differ:

\begin{itemize}
\tightlist
\item
  \textbf{Visibility:} The contracting agency may have limited
  visibility into whether a contractor is using agentic tools, what
  proportion of deliverables are agent-generated, and whether the
  contractor's review processes address the failure modes in Appendix A.
  ISM-2074's AI usage policy requirement applies to the agency's own
  use; it is less clear how it flows down to contracted development.
\item
  \textbf{Acceptance criteria:} Current delivery acceptance criteria for
  contracted software development typically focus on functional
  requirements, test coverage, and compliance with coding standards.
  They do not typically address the semantic correctness properties this
  paper identifies --- trust boundary maintenance, audit trail
  integrity, context-appropriate error handling. A contractor could
  deliver code that meets every contractual requirement while containing
  systematic ACF-pattern violations.
\item
  \textbf{Review responsibility:} When a contractor delivers
  agent-generated code, who is responsible for the security-focused
  review --- the contractor's internal review process, the agency's
  acceptance review, or both? If the agency relies on the contractor's
  review, the agency inherits the contractor's review capacity
  constraints and habituation dynamics (Section 4.2).
\item
  \textbf{Correlated risk across contracts:} If multiple agencies
  contract the same provider, and that provider uses the same agent
  tooling and prompts across engagements, the correlated failure problem
  (Section 9.7) extends across agency boundaries through the contractor,
  even if the agencies have no direct relationship.
\end{itemize}

This is arguably the most operationally significant gap for ASD to
address, because it is where the majority of government code is actually
produced. However, the contracted development assumption itself may be
shifting. Agentic AI lowers the barrier to code production --- not just
for in-house development teams, but for the non-developer power users
described in Section 1.2.8. If agencies begin generating more code
internally, whether through technical staff using agents or through
analysts and operators producing executable logic, the balance between
contracted and in-house code may change. The governance challenge then
compounds: agencies must address agentic code risks in both their
contracted deliverables and their own expanding internal code
production. (See Recommendation 6, §10.1, for candidate controls.)

\hypertarget{recommendations}{%
\section{10. Recommendations}\label{recommendations}}

The following recommendations are \textbf{candidate controls presented
for community consultation}, not normative guidance. They reflect the
authors' assessment of what is achievable and useful, grounded in the
observed patterns and analytical framework in this paper. They are
grouped by audience. For organisations seeking immediate priorities,
three recommendations have the highest leverage and are achievable in
the near term:

\begin{itemize}
\tightlist
\item
  \textbf{Recommendation 2} (§10.1, for policy bodies): \emph{Issue
  guidance on treating agent output as a trust boundary.} This provides
  the conceptual foundation all other controls build on.
\item
  \textbf{Recommendation 3} (§10.1, for policy bodies): \emph{Extend ISM
  controls for agent-generated code.} The June 2025 ISM overhaul
  provides strong foundations that need only targeted extensions.
\item
  \textbf{Recommendation 11} (§10.3, for organisations): \emph{Document
  institutional security knowledge in machine-readable form.} This is
  the most direct defence against agents that lack institutional
  context, and requires no new tooling.
\end{itemize}

\hypertarget{for-security-policy-bodies-asd-acsc}{%
\subsection{10.1 For Security Policy Bodies (ASD,
ACSC)}\label{for-security-policy-bodies-asd-acsc}}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  \textbf{Develop a taxonomy of agentic code failure modes} based on
  established frameworks (STRIDE or equivalent). This paper's taxonomy
  (Appendix A) is a starting point, not a final product. The taxonomy
  should be developed in consultation with organisations that are
  already using agentic coding in assessed systems, to ground it in
  observed failure modes rather than theoretical concerns.
\item
  \textbf{Issue guidance on treating agent output as a trust boundary.}
  Clarify how agent-generated code should be classified in the trust
  model and what validation is required before integration into assessed
  systems. At minimum, guidance should address:

  \begin{itemize}
  \tightlist
  \item
    Whether agent-generated code requires different review criteria than
    human-authored code
  \item
    Whether organisations must track which code was agent-generated
    (provenance)
  \item
    What constitutes ``sufficient validation'' at the agent output
    boundary
  \end{itemize}
\item
  \textbf{Extend ISM controls for software development} to address the
  distinct risk profile of agent-generated code. The June 2025 ISM
  overhaul added strong foundations (ISM-2060/2061 for code review,
  ISM-0402 for comprehensive testing, ISM-2026--2028 for artefact
  integrity), but these controls assume human-paced development.
  Candidate extensions:

  \begin{itemize}
  \tightlist
  \item
    ISM-0401 (Secure by Design): Require that Secure by Design practices
    include machine-enforceable rules when AI agents are used, since
    agents cannot internalise design principles from documentation.
  \item
    ISM-2060/2061 (code review / security-focused peer review): Address
    review capacity scaling and the habituation effect under
    agent-volume code generation. Consider requiring measured review
    effectiveness, not just documented process.
  \item
    ISM-0402 (testing with SAST/DAST/SCA): Extend SAST requirements to
    include semantic boundary analysis --- trust tier flow verification,
    defensive pattern detection in audit-critical contexts --- beyond
    current known-vulnerability-pattern scanning.
  \item
    ISM-2026--2028 (artefact integrity): Extend supply chain controls to
    cover agent-generated first-party code as a supply chain input with
    distinct risk properties (correlated failures, no persistent
    learning).
  \item
    New control: Require organisations using agentic coding to
    demonstrate that their review and validation processes remain
    effective under the volume of agent-generated code they produce.
  \end{itemize}
\item
  \textbf{Commission research into automated semantic boundary
  enforcement} as a complementary control to human review. The technical
  feasibility finding (Appendix B) suggests this is tractable --- the
  core analysis uses only Python's standard library AST module and the
  verification properties are independently auditable (Section B.5).
  Research priorities:

  \begin{itemize}
  \tightlist
  \item
    Validate the feasibility finding against a broader set of codebases
    and programming languages
  \item
    Evaluate whether automated enforcement meaningfully reduces defect
    escape rates in practice
  \item
    Develop criteria for assessing the assurance properties of
    enforcement tools themselves (Appendix B, Section B.5)
  \end{itemize}
\item
  \textbf{Extend SDLC-equivalent controls to executable logic produced
  outside formal development teams.} Agentic tools are expanding the
  population of software producers beyond professional developers
  (Section 1.2.8). Organisations should inventory and govern
  agent-generated plugins, automations, BI extensions, workflow scripts,
  low-code components, and similar artefacts produced by analysts,
  operators, and other non-developer staff. Where such artefacts affect
  trust boundaries, access control, audit trails, or data integrity,
  they should be subject to provenance, review, and validation controls
  proportionate to their impact --- even when the producers do not
  consider themselves developers and the artefacts do not live in a
  formal version control system. The governance perimeter has expanded;
  controls must follow it.
\item
  \textbf{Address agent-generated code in contracted development.} The
  majority of software development for Australian Government systems is
  performed by contracted service providers (Section 9.8). Contracted
  developers' agents produce the same correlated, context-inappropriate
  patterns as in-house agents, but with additional governance
  challenges. Candidate controls:

  \begin{itemize}
  \tightlist
  \item
    Contract clauses requiring disclosure of agentic tool use and
    provenance tracking for agent-generated deliverables
  \item
    Acceptance criteria that include semantic correctness properties
    (trust boundary maintenance, audit trail integrity,
    context-appropriate error handling), not just functional
    requirements and test coverage
  \item
    Clarification of how ISM-2074's AI usage policy requirement flows
    down to contracted development
  \item
    Assessment of correlated failure risk when the same contractor and
    agent tooling serves multiple agencies --- a systematic defect
    introduced by a contractor's agent may propagate across agency
    boundaries through the shared provider
  \end{itemize}
\end{enumerate}

\hypertarget{for-irap-assessors}{%
\subsection{10.2 For IRAP Assessors}\label{for-irap-assessors}}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  \textbf{Include agentic development practices in assessment scope}
  when organisations use AI coding agents in assessed systems. Evaluate:

  \begin{itemize}
  \tightlist
  \item
    The validation boundary between agent output and code integration
    (Section 5.3)
  \item
    Evidence of review effectiveness under volume, not just existence of
    review process
  \item
    Provenance tracking for agent-generated code
  \item
    Whether the organisation has identified and addressed the failure
    modes relevant to their system's classification and data sensitivity
  \end{itemize}
\item
  \textbf{Assess review quality, not just review process.} A documented
  code review process that is overwhelmed by volume provides less
  assurance than a smaller, more thorough review supported by automated
  pre-screening. Ask for evidence: defect escape rates, review depth
  audits, or automated pre-screening coverage metrics (Section 9.4).
\item
  \textbf{Consider correlated failure risk.} Traditional defect models
  assume independent failures. Agent-generated code produces correlated
  failures (Section 9.7). When assessing test coverage and defect rates,
  verify that the organisation's testing strategy accounts for
  correlation --- finding one instance of a failure pattern should
  trigger codebase-wide scanning for the same pattern.
\end{enumerate}

\hypertarget{for-organisations-using-agentic-coding}{%
\subsection{10.3 For Organisations Using Agentic
Coding}\label{for-organisations-using-agentic-coding}}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  \textbf{Treat agent-generated code as external input} requiring
  validation at the boundary. Do not assume agent output is correct
  because it passes tests and follows conventions. The failure modes in
  Appendix A are specifically designed to pass tests and look correct
  --- that is what makes them dangerous.
\item
  \textbf{Document your institutional security knowledge in
  machine-readable form.} The gap between ``what Python permits'' and
  ``what our system requires'' is institutional knowledge that currently
  lives in documentation, team culture, and individual expertise.
  Encoding it in machine-checkable rules --- whether through a
  purpose-built tool (Appendix B), project-specific linter rules, or
  structured review checklists --- is the most direct defence against
  agents that don't share that knowledge.
\item
  \textbf{Invest in automated semantic boundary enforcement} as a
  complement to human review. The human review budget is finite;
  automated tools that handle structural trust boundary violations free
  human reviewers to focus on semantic issues that require institutional
  knowledge. The case study (Section 8.5) demonstrates that this is a
  redirection of existing effort, not additional overhead.
\item
  \textbf{Measure and monitor review effectiveness} under agentic
  volume. If review quality is degrading, address it --- through
  tooling, capacity, or rate-limiting --- before it becomes a compliance
  gap. Don't wait for an incident to discover that the review process is
  no longer providing the assurance it claims.
\item
  \textbf{Develop domain-specific review prompts aligned to your threat
  model.} Rather than relying on generic ``review this code''
  instructions, create prompted analytical perspectives that map to
  specific failure categories --- security boundary review for
  ACF-T1/E1, operational resilience review for ACF-R1/R2, data quality
  review for ACF-S1/S2. Critically, do not limit these to technical
  perspectives. Analytical frames like systems thinking (detecting
  feedback loops, shifted burdens, and second-order effects) and quality
  engineering (evaluating test strategy, coverage gaps, and verification
  adequacy) catch classes of defect that purely technical prompts miss
  --- organisations will naturally think of the technical skills they
  need but overlook the logical and analytical ones. These prompts are
  reviewable, version-controlled security artefacts that encode
  institutional knowledge in a form complementary to the
  machine-readable rules described in Section 7.2 and Appendix B. Run
  multiple prompted perspectives as a structured pre-screening step
  before human review to increase coverage breadth at agent speed
  (Section 7.1).
\item
  \textbf{When an agent-generated defect is found, scan for the pattern,
  not just the instance.} Correlated failures mean the same defect
  likely exists in other agent-generated code. Treating each defect as
  isolated underestimates the actual risk.
\item
  \textbf{Contribute to the community vocabulary.} Document your
  organisation's experience with agentic code failure modes and share
  (at an appropriate classification) to build the collective
  understanding. The taxonomy in this paper was developed from a single
  project's experience; it needs validation and extension from diverse
  government contexts.
\end{enumerate}

\hypertarget{appendix-a-agentic-code-failure-taxonomy}{%
\section{Appendix A: Agentic Code Failure
Taxonomy}\label{appendix-a-agentic-code-failure-taxonomy}}

A structured catalogue of failure modes, mapped to STRIDE categories,
with detection characteristics, code examples, and risk ratings. Each
entry includes the \emph{reason agents produce this pattern} ---
understanding why helps calibrate both detection tools and review
processes.

\hypertarget{summary-table}{%
\subsection{Summary Table}\label{summary-table}}

\begin{longtable}[]{@{}lllll@{}}
\toprule\noalign{}
ID & Name & STRIDE & Risk & Existing Detection \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
ACF-S1 & Competence spoofing & Spoofing & High & None \\
ACF-S2 & Hallucinated field access & Spoofing & High & Partial \\
ACF-S3 & Structural identity spoofing & Spoofing & High & Partial \\
ACF-T1 & Trust tier conflation & Tampering & Critical & None \\
ACF-T2 & Silent coercion & Tampering & Medium & None \\
ACF-R1 & Audit trail destruction & Repudiation & High & Partial \\
ACF-R2 & Partial completion & Repudiation & High & None \\
ACF-I1 & Verbose error response & Info Disclosure & Medium & Partial \\
ACF-I2 & Stack trace exposure & Info Disclosure & Low & Good \\
ACF-D1 & Finding flood & DoS & High & N/A \\
ACF-D2 & Review capacity exhaustion & DoS & High & N/A \\
ACF-E1 & Implicit privilege grant & Elevation & Critical & None \\
ACF-E2 & Unvalidated delegation & Elevation & High & Partial \\
\end{longtable}

\hypertarget{detailed-entries}{%
\subsection{Detailed Entries}\label{detailed-entries}}

\emph{Policy readers: the Summary Table above and the Detection
Capability Summary at the end of this appendix provide a complete
overview without requiring code fluency. The detailed entries below,
which include Python code examples, are provided for technical
practitioners, tool builders, and organisations implementing detection
capabilities. Non-Python readers may rely on the Description, Why it's
dangerous, and Detection approach fields for each entry.}

\textbf{Language specificity.} The code examples throughout this
appendix use Python, reflecting the case study environment. The failure
modes vary in language-generality:

\begin{itemize}
\tightlist
\item
  \textbf{Language-general} (applicable across Python, Java, C\#,
  TypeScript, Go, etc.): ACF-T1 (trust tier conflation), ACF-T2 (silent
  coercion), ACF-R1 (audit trail destruction), ACF-R2 (partial
  completion), ACF-I1 (verbose error response), ACF-D1 (finding flood),
  ACF-D2 (review capacity exhaustion), ACF-E1 (implicit privilege
  grant), ACF-E2 (unvalidated delegation). The failure \emph{patterns}
  differ by language (e.g., \texttt{catch\ (Exception\ e)} in Java,
  \texttt{catch} in C++, \texttt{recover()} in Go), but the failure
  \emph{mode} is the same.
\item
  \textbf{Python-specific surface form} (same underlying failure,
  different manifestation in other languages): ACF-S1 (\texttt{.get()}
  with defaults --- other languages have analogues like
  \texttt{Optional.orElse()} in Java or \texttt{??} in C\#), ACF-S2
  (\texttt{getattr()} with defaults --- Python-specific, though dynamic
  languages like Ruby have \texttt{send}/\texttt{respond\_to?}), ACF-S3
  (\texttt{hasattr()} as capability gate --- Python-specific surface
  form, though the underlying failure applies to any language with duck
  typing or structural typing; Ruby's \texttt{respond\_to?}, Go's
  interface satisfaction, and TypeScript's structural type compatibility
  are analogues), ACF-I2 (stack trace exposure --- the specific
  mechanisms are framework-dependent everywhere).
\end{itemize}

Organisations working in other languages should read the
\emph{Description} and \emph{Why it's dangerous} fields as
language-general, and treat the \emph{Example} and \emph{Detection
approach} fields as Python-specific reference implementations.

\hypertarget{acf-s1-competence-spoofing}{%
\subsubsection{ACF-S1: Competence
Spoofing}\label{acf-s1-competence-spoofing}}

\textbf{STRIDE:} Spoofing \textbar{} \textbf{Risk:} High \textbar{}
\textbf{Detection:} None

\textbf{Description:} Default values fabricate data where the absence of
data should be surfaced as a failure, error, or explicit ``unknown.''
The code presents a confident result that is actually based on
fabricated input.

\textbf{Why agents produce this:} The \texttt{.get(key,\ default)}
pattern appears in millions of Python files. In most contexts, providing
a default for missing keys is genuinely good practice --- a web
application displaying ``Unknown'' for a missing user name is fine.
Agents learn this as a universal pattern and apply it in contexts where
the default fabricates safety-critical data.

\textbf{Example:}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# Agent{-}generated — looks defensive and robust}
\KeywordTok{def}\NormalTok{ assess\_risk\_level(record):}
\NormalTok{    classification }\OperatorTok{=}\NormalTok{ record.get(}\StringTok{"security\_classification"}\NormalTok{, }\StringTok{"OFFICIAL"}\NormalTok{)}
\NormalTok{    clearance }\OperatorTok{=}\NormalTok{ record.get(}\StringTok{"required\_clearance"}\NormalTok{, }\StringTok{"baseline"}\NormalTok{)}
    \ControlFlowTok{return}\NormalTok{ classification, clearance}

\CommentTok{\# Correct for high{-}assurance context — absence is a failure}
\KeywordTok{def}\NormalTok{ assess\_risk\_level(record):}
    \ControlFlowTok{if} \StringTok{"security\_classification"} \KeywordTok{not} \KeywordTok{in}\NormalTok{ record:}
        \ControlFlowTok{raise}\NormalTok{ MissingSecurityClassification(}
            \SpecialStringTok{f"Record }\SpecialCharTok{\{}\NormalTok{record[}\StringTok{\textquotesingle{}id\textquotesingle{}}\NormalTok{]}\SpecialCharTok{\}}\SpecialStringTok{: security\_classification absent — "}
            \SpecialStringTok{f"upstream data integrity failure, cannot assess risk"}
\NormalTok{        )}
    \ControlFlowTok{if} \StringTok{"required\_clearance"} \KeywordTok{not} \KeywordTok{in}\NormalTok{ record:}
        \ControlFlowTok{raise}\NormalTok{ MissingSecurityClearance(}
            \SpecialStringTok{f"Record }\SpecialCharTok{\{}\NormalTok{record[}\StringTok{\textquotesingle{}id\textquotesingle{}}\NormalTok{]}\SpecialCharTok{\}}\SpecialStringTok{: required\_clearance absent — "}
            \SpecialStringTok{f"cannot determine access level, refusing to default"}
\NormalTok{        )}
    \ControlFlowTok{return}\NormalTok{ record[}\StringTok{"security\_classification"}\NormalTok{], record[}\StringTok{"required\_clearance"}\NormalTok{]}
\end{Highlighting}
\end{Shaded}

\textbf{Why it's dangerous:} The first version silently downgrades
security classifications when data is missing. A PROTECTED document with
a corrupted or missing \texttt{security\_classification} field is
treated as OFFICIAL. Downstream access control decisions are based on
the fabricated classification.

\textbf{Detection approach:} Flag \texttt{.get()} and \texttt{getattr()}
with defaults on objects whose type is annotated with a trust tier of
Tier 1 (internal/audit) or Tier 2 (validated pipeline data). Requires
trust tier annotations (not available in existing tools).

\hypertarget{acf-s2-hallucinated-field-access}{%
\subsubsection{ACF-S2: Hallucinated Field
Access}\label{acf-s2-hallucinated-field-access}}

\textbf{STRIDE:} Spoofing \textbar{} \textbf{Risk:} High \textbar{}
\textbf{Detection:} Partial

\textbf{Description:} Agent accesses a field name that doesn't exist on
the target object, masked by \texttt{getattr()} with a default. The code
operates on fabricated data while appearing to access a real field.

\textbf{Why agents produce this:} Agents occasionally hallucinate field
names --- predicting a plausible field name that doesn't exist in the
actual schema. Without \texttt{getattr}, this produces an immediate
\texttt{AttributeError}. With
\texttt{getattr(obj,\ "hallucinated\_field",\ None)}, the error is
silently suppressed and the code operates on \texttt{None} (or whatever
default is provided).

\textbf{Example:}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# Agent hallucinated "risk\_score" — actual field is "risk\_rating"}
\NormalTok{threshold }\OperatorTok{=} \BuiltInTok{getattr}\NormalTok{(assessment, }\StringTok{"risk\_score"}\NormalTok{, }\DecValTok{0}\NormalTok{)}
\ControlFlowTok{if}\NormalTok{ threshold }\OperatorTok{\textgreater{}} \DecValTok{5}\NormalTok{:}
\NormalTok{    escalate(assessment)}
\CommentTok{\# risk\_score is always 0 (the default), so nothing is ever escalated.}
\CommentTok{\# The code looks correct. Tests pass (they test the escalation path with explicit values).}
\CommentTok{\# The bug is invisible until someone notices that escalation never triggers.}
\end{Highlighting}
\end{Shaded}

\textbf{Why it's dangerous:} The code silently does nothing instead of
crashing. In a security context, ``nothing happens'' can mean ``threats
are not escalated'' or ``alerts are not raised'' --- failures of
omission that are harder to detect than failures of commission.

\textbf{Detection approach:} Type checkers (mypy, pyright) catch this
\emph{if the object is fully annotated}. If the object is \texttt{Any}
or untyped, type checkers are silent. The semantic boundary enforcer
(Appendix B) adds a complementary rule: \texttt{getattr} with a default
on any object that has a declared type annotation is flagged, because
the annotation means the field set is known and access should be direct.

\hypertarget{acf-s3-structural-identity-spoofing}{%
\subsubsection{ACF-S3: Structural Identity
Spoofing}\label{acf-s3-structural-identity-spoofing}}

\textbf{STRIDE:} Spoofing (+ Elevation of Privilege consequence)
\textbar{} \textbf{Risk:} High \textbar{} \textbf{Detection:} Partial

\textbf{Description:} A \texttt{hasattr()} check is used as a capability
or privilege gate, allowing any object that declares the expected
attribute to pass --- regardless of whether the object is of the correct
type. The gate accepts structural presence as proof of identity.

\textbf{Why agents produce this:} \texttt{hasattr()} is the idiomatic
Python pattern for duck-typing capability checks. Training data is
saturated with it --- agents building plugin systems, authorisation
checks, or capability dispatchers will reach for \texttt{hasattr} by
default because it is the ``Pythonic'' way to test whether an object
supports an operation. The concept that structural presence is not
ontological identity --- that \emph{having} an attribute is not the same
as \emph{being} the right type --- is a security distinction that the
language actively discourages.

\textbf{Example:}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# Agent{-}generated — "Pythonic" duck{-}typing capability check}
\KeywordTok{def}\NormalTok{ process\_classified(obj):}
    \ControlFlowTok{if} \BuiltInTok{hasattr}\NormalTok{(obj, }\StringTok{"security\_clearance"}\NormalTok{):}
\NormalTok{        handle\_classified(obj)  }\CommentTok{\# Any object with this attr gets in}

\CommentTok{\# Trivial bypass — no type hierarchy modification needed}
\KeywordTok{class}\NormalTok{ Impersonator:}
\NormalTok{    security\_clearance }\OperatorTok{=} \StringTok{"TOP\_SECRET"}  \CommentTok{\# Just declare the attribute}

\NormalTok{process\_classified(Impersonator())  }\CommentTok{\# Gate opens}

\CommentTok{\# Correct — requires actual type membership}
\KeywordTok{def}\NormalTok{ process\_classified(obj):}
    \ControlFlowTok{if} \BuiltInTok{isinstance}\NormalTok{(obj, ClearedPersonnel):}
\NormalTok{        handle\_classified(obj)  }\CommentTok{\# Must inherit from ClearedPersonnel}
    \CommentTok{\# Cannot bypass without modifying the class hierarchy itself}
\end{Highlighting}
\end{Shaded}

\textbf{Why it's dangerous:} Unlike ACF-S1 (data fabrication via
defaults) where the fabricated value is visible at the call site, the
exploit surface for \texttt{hasattr} gates is anywhere an object is
constructed --- potentially far from the gate. The gate looks secure in
isolation. Worse, Python's \texttt{\_\_getattr\_\_} protocol means a
single class can dynamically claim to possess \emph{any} attribute:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{class}\NormalTok{ UniversalImpersonator:}
    \KeywordTok{def} \FunctionTok{\_\_getattr\_\_}\NormalTok{(}\VariableTok{self}\NormalTok{, name):}
        \ControlFlowTok{return} \VariableTok{True}  \CommentTok{\# "Yes, I have that. And everything else."}

\CommentTok{\# This object passes EVERY hasattr check in the entire codebase.}
\CommentTok{\# An isinstance check is immune to this.}
\end{Highlighting}
\end{Shaded}

This is the capability-based equivalent of ACF-S1's competence spoofing:
ACF-S1 fabricates \emph{data} where absence should be a failure; ACF-S3
fabricates \emph{identity} where type membership should be required. The
object claims to be something it isn't, and the gate believes it because
the check is structural (has the attribute) rather than ontological (is
the type). The elevation of privilege consequence follows directly ---
the impersonator passes through a privilege gate that should have
rejected it.

\textbf{Detection approach:} An unconditional lint rule banning
\texttt{hasattr()} catches all instances (the case study codebase in
Section 8 enforces this). General-purpose linters do not flag
\texttt{hasattr} because it is considered idiomatic Python. The semantic
boundary enforcer (Appendix B) treats \texttt{hasattr} as
unconditionally prohibited --- unlike \texttt{.get()} or
\texttt{getattr()}, which are context-dependent, there is no legitimate
use of \texttt{hasattr} that cannot be expressed more safely as
\texttt{isinstance()}, explicit
\texttt{try}/\texttt{except\ AttributeError}, or an allowset check.
Detection is rated Partial because the rule is simple to implement but
not present in any widely-deployed tool.

\hypertarget{acf-t1-trust-tier-conflation}{%
\subsubsection{ACF-T1: Trust Tier
Conflation}\label{acf-t1-trust-tier-conflation}}

\textbf{STRIDE:} Tampering \textbar{} \textbf{Risk:} Critical \textbar{}
\textbf{Detection:} None

\textbf{Description:} Data from an external (untrusted) source is used
in an internal (trusted) context without passing through a validation
boundary. The data's effective trust level is silently elevated.

\textbf{Why agents produce this:} Python's type system doesn't
distinguish between data from different sources. A \texttt{dict} from
\texttt{requests.get().json()} and a \texttt{dict} from a validated
internal query are the same type. Agents see both as ``a dict'' and
treat them interchangeably because nothing in the language tells them
otherwise.

\textbf{Example:}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# Agent{-}generated — clean, readable, catastrophically wrong}
\KeywordTok{def}\NormalTok{ sync\_partner\_records(partner\_api\_url):}
\NormalTok{    response }\OperatorTok{=}\NormalTok{ requests.get(}\SpecialStringTok{f"}\SpecialCharTok{\{}\NormalTok{partner\_api\_url}\SpecialCharTok{\}}\SpecialStringTok{/records"}\NormalTok{)}
\NormalTok{    records }\OperatorTok{=}\NormalTok{ response.json()}
    \ControlFlowTok{for}\NormalTok{ record }\KeywordTok{in}\NormalTok{ records:}
\NormalTok{        db.execute(}
\NormalTok{            insert(internal\_records).values(}\OperatorTok{**}\NormalTok{record)}
\NormalTok{        )}
    \CommentTok{\# External data inserted directly into internal database.}
    \CommentTok{\# No schema validation, no field allowlisting, no type checking.}
    \CommentTok{\# Partner could send arbitrary fields, wrong types, injection payloads.}

\CommentTok{\# Correct — validate at the boundary}
\KeywordTok{def}\NormalTok{ sync\_partner\_records(partner\_api\_url):}
\NormalTok{    response }\OperatorTok{=}\NormalTok{ requests.get(}\SpecialStringTok{f"}\SpecialCharTok{\{}\NormalTok{partner\_api\_url}\SpecialCharTok{\}}\SpecialStringTok{/records"}\NormalTok{)}
\NormalTok{    raw\_records }\OperatorTok{=}\NormalTok{ response.json()}
    \ControlFlowTok{for}\NormalTok{ raw }\KeywordTok{in}\NormalTok{ raw\_records:}
        \ControlFlowTok{try}\NormalTok{:}
\NormalTok{            validated }\OperatorTok{=}\NormalTok{ PartnerRecordSchema.validate(raw)}
        \ControlFlowTok{except}\NormalTok{ ValidationError }\ImportTok{as}\NormalTok{ e:}
\NormalTok{            quarantine(raw, reason}\OperatorTok{=}\BuiltInTok{str}\NormalTok{(e))}
            \ControlFlowTok{continue}
\NormalTok{        db.execute(}
\NormalTok{            insert(internal\_records).values(}
\NormalTok{                name}\OperatorTok{=}\NormalTok{validated.name,}
\NormalTok{                status}\OperatorTok{=}\NormalTok{validated.status,}
\NormalTok{            )}
\NormalTok{        )}
\end{Highlighting}
\end{Shaded}

\textbf{Why it's dangerous:} This is the most critical failure mode
because it compromises the integrity of the internal data store --- the
system's source of truth. Once external data enters the internal store
without validation, every downstream consumer trusts it as internal
data. Corruption propagates invisibly.

\textbf{Detection approach:} Taint analysis --- trace the return values
of functions marked \texttt{@external\_boundary} (or matched by the
known external call heuristic list) and flag if they reach data store
operations without passing through a function marked
\texttt{@validates\_external}. This is the core capability of the tool
described in Appendix B.

\hypertarget{acf-t2-silent-coercion}{%
\subsubsection{ACF-T2: Silent Coercion}\label{acf-t2-silent-coercion}}

\textbf{STRIDE:} Tampering \textbar{} \textbf{Risk:} Medium \textbar{}
\textbf{Detection:} None

\textbf{Description:} Type coercion across trust boundaries hides data
quality issues. Values are silently converted to a compatible type
rather than being flagged as invalid.

\textbf{Why agents produce this:} Python's \texttt{or} operator and
conditional expressions make coercion easy and idiomatic.
\texttt{value\ =\ input\_value\ or\ "default"} is a common pattern.
Agents apply it broadly without distinguishing between contexts where
coercion is appropriate (Tier 3 → Tier 2 at a validated boundary) and
contexts where it's dangerous (Tier 1 internal data that should never
need coercion).

\textbf{Example:}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# Silent coercion hides data quality problem}
\NormalTok{amount }\OperatorTok{=} \BuiltInTok{float}\NormalTok{(row.get(}\StringTok{"transaction\_amount"}\NormalTok{, }\DecValTok{0}\NormalTok{))}
\CommentTok{\# Missing transaction amount is silently zero — not "unknown" or "error."}
\CommentTok{\# A zero{-}value transaction passes every downstream check.}
\CommentTok{\# An audit query for "transactions over $1000" won\textquotesingle{}t find it,}
\CommentTok{\# but neither will "transactions with missing amounts."}

\CommentTok{\# Better — make the absence explicit}
\ControlFlowTok{if} \StringTok{"transaction\_amount"} \KeywordTok{not} \KeywordTok{in}\NormalTok{ row:}
    \ControlFlowTok{return}\NormalTok{ TransformResult.error(\{}\StringTok{"reason"}\NormalTok{: }\StringTok{"missing\_amount"}\NormalTok{, }\StringTok{"row\_id"}\NormalTok{: row\_id\})}
\NormalTok{amount }\OperatorTok{=} \BuiltInTok{float}\NormalTok{(row[}\StringTok{"transaction\_amount"}\NormalTok{])}
\end{Highlighting}
\end{Shaded}

\textbf{Detection approach:} Flag coercion patterns (\texttt{.get()}
with non-None defaults, \texttt{or} chains with fallback values, ternary
expressions with defaults) on fields from Tier 1 or Tier 2 data. The
distinction from ACF-S1 is that this involves type conversion, not just
default substitution.

\hypertarget{acf-r1-audit-trail-destruction}{%
\subsubsection{ACF-R1: Audit Trail
Destruction}\label{acf-r1-audit-trail-destruction}}

\textbf{STRIDE:} Repudiation \textbar{} \textbf{Risk:} High \textbar{}
\textbf{Detection:} Partial

\textbf{Description:} Broad exception handlers catch errors from
audit-critical operations and log-and-continue rather than propagating
the failure to the audit system.

\textbf{Why agents produce this:} ``Catch exceptions and log them'' is a
pervasive pattern in training data. In most applications, it's
reasonable --- a web server should log errors and keep serving. Agents
apply this pattern to audit-critical operations without recognising that
some failures must propagate rather than be absorbed.

\textbf{Example:}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# Agent{-}generated — looks like responsible error handling}
\ControlFlowTok{try}\NormalTok{:}
\NormalTok{    record\_decision(case\_id, decision, rationale, evidence)}
\ControlFlowTok{except} \PreprocessorTok{Exception} \ImportTok{as}\NormalTok{ e:}
\NormalTok{    logger.error(}\SpecialStringTok{f"Failed to record decision for }\SpecialCharTok{\{}\NormalTok{case\_id}\SpecialCharTok{\}}\SpecialStringTok{: }\SpecialCharTok{\{}\NormalTok{e}\SpecialCharTok{\}}\SpecialStringTok{"}\NormalTok{)}
    \CommentTok{\# Decision was made. Decision was not recorded.}
    \CommentTok{\# The audit trail now has a gap that cannot be reconstructed.}
    \CommentTok{\# The log message may be rotated away. The decision stands, unrecorded.}
\end{Highlighting}
\end{Shaded}

\textbf{Why it's dangerous:} In regulatory contexts, the audit trail is
the legal record. A gap in the audit trail is not just a logging failure
--- it's a compliance failure that may have legal consequences. ``We
made a decision but can't prove what it was based on'' is an
unacceptable answer in a formal inquiry.

\textbf{Detection approach:} Existing linters flag bare \texttt{except:}
(no exception type) but not \texttt{except\ Exception:} (which is
considered acceptable practice). Semantic detection requires
understanding which operations are audit-critical --- this is
project-specific knowledge encoded in the trust topology (e.g.,
functions annotated as audit-write operations should not be inside broad
exception handlers that continue on failure).

\hypertarget{acf-r2-partial-completion}{%
\subsubsection{ACF-R2: Partial
Completion}\label{acf-r2-partial-completion}}

\textbf{STRIDE:} Repudiation \textbar{} \textbf{Risk:} High \textbar{}
\textbf{Detection:} None

\textbf{Description:} A sequence of operations that should be atomic
(all-or-nothing) is implemented without rollback, so partial failure
leaves the system in an inconsistent state.

\textbf{Why agents produce this:} Agents implement operations
sequentially and add error handling per-step. They don't naturally
recognise that a group of operations should be treated as a transaction
unless explicitly prompted. The concept of ``these three operations must
all succeed or all fail'' is a design decision, not a language feature.

\textbf{Example:}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# Agent{-}generated — each step has error handling, but no atomicity}
\KeywordTok{def}\NormalTok{ reclassify\_document(doc\_id, new\_classification):}
\NormalTok{    update\_classification(doc\_id, new\_classification)  }\CommentTok{\# Step 1: succeeds}
\NormalTok{    notify\_stakeholders(doc\_id, new\_classification)     }\CommentTok{\# Step 2: fails (network error)}
\NormalTok{    record\_reclassification(doc\_id, old, new)           }\CommentTok{\# Step 3: never runs}
    \CommentTok{\# Document is reclassified, stakeholders don\textquotesingle{}t know, audit trail is incomplete.}
    \CommentTok{\# If step 2 is wrapped in try/except and continues, step 3 records a}
    \CommentTok{\# reclassification that stakeholders were never notified about.}
\end{Highlighting}
\end{Shaded}

\textbf{Detection approach:} No existing tool detects this --- it
requires understanding which operations form a logical transaction. A
semantic boundary enforcer could flag functions that contain multiple
audit-write operations without a transaction context, but this requires
project-specific annotation of which operations are audit-critical.

\hypertarget{acf-i1-verbose-error-response}{%
\subsubsection{ACF-I1: Verbose Error
Response}\label{acf-i1-verbose-error-response}}

\textbf{STRIDE:} Information Disclosure \textbar{} \textbf{Risk:} Medium
\textbar{} \textbf{Detection:} Partial

\textbf{Description:} Error handlers expose internal system details
(database schemas, file paths, query parameters, library versions) in
error responses.

\textbf{Why agents produce this:} Agents produce ``helpful'' error
messages that include full context. During development, this is
valuable. In production, it's reconnaissance information. Agents don't
distinguish between development and production error handling because
the distinction is contextual, not syntactic.

\textbf{Example:}

\begin{Shaded}
\begin{Highlighting}[]
\ControlFlowTok{except}\NormalTok{ DatabaseError }\ImportTok{as}\NormalTok{ e:}
    \ControlFlowTok{return}\NormalTok{ \{}\StringTok{"error"}\NormalTok{: }\BuiltInTok{str}\NormalTok{(e), }\StringTok{"query"}\NormalTok{: sql, }\StringTok{"connection"}\NormalTok{: }\BuiltInTok{str}\NormalTok{(db\_url)\}}
\end{Highlighting}
\end{Shaded}

\textbf{Detection approach:} Existing scanners detect some cases
(credential patterns, known sensitive variable names). Comprehensive
detection requires understanding which variables contain sensitive
information --- a context-dependent judgment.

\hypertarget{acf-i2-stack-trace-exposure}{%
\subsubsection{ACF-I2: Stack Trace
Exposure}\label{acf-i2-stack-trace-exposure}}

\textbf{STRIDE:} Information Disclosure \textbar{} \textbf{Risk:} Low
\textbar{} \textbf{Detection:} Good

\emph{Note: This entry is included for taxonomy completeness. It is
well-covered by existing tooling and is lower risk than other entries.
It could reasonably be treated as a sub-variant of ACF-I1.}

\textbf{Description:} Full Python tracebacks returned in API responses
or user-facing error messages.

\textbf{Why agents produce this:} \texttt{traceback.format\_exc()} is a
common debugging pattern. Agents include it in error handlers without
considering the deployment context.

\textbf{Detection approach:} Well-covered by existing tools. Most
security scanners and web framework linters detect traceback exposure.

\hypertarget{acf-d1-finding-flood}{%
\subsubsection{ACF-D1: Finding Flood}\label{acf-d1-finding-flood}}

\textbf{STRIDE:} Denial of Service \textbar{} \textbf{Risk:} High
\textbar{} \textbf{Detection:} N/A (process threat)

\textbf{Description:} The volume of static analysis findings on
agent-generated code overwhelms reviewers, causing them to rubber-stamp
findings without evaluation.

\textbf{Why this happens:} Agents produce code at volume, and if that
code triggers many findings, the review queue grows faster than the
review capacity. Reviewers under volume pressure shift from evaluating
each finding to batch-dismissing them. The DoS is against the
\emph{review process}, not the system.

\textbf{Mitigation:} Finding caps per rule per file (Appendix B),
prioritised finding presentation (critical first), and measured
suppression rates as a health metric.

\hypertarget{acf-d2-review-capacity-exhaustion}{%
\subsubsection{ACF-D2: Review Capacity
Exhaustion}\label{acf-d2-review-capacity-exhaustion}}

\textbf{STRIDE:} Denial of Service \textbar{} \textbf{Risk:} High
\textbar{} \textbf{Detection:} N/A (process threat)

\textbf{Description:} Agent code generation velocity exceeds the
organisation's capacity for security-focused review, degrading review
from active verification to passive scanning.

\textbf{Why this happens:} Agents can generate plausible,
convention-conforming code faster than review processes were designed to
absorb (Section 1.2.1). Review capacity doesn't scale at the same rate.
The review process becomes a bottleneck, and the organisational response
is often to lower the review bar rather than reduce the generation rate.

\textbf{Mitigation:} Automated pre-screening to reduce the human review
burden, volume-aware capacity planning, and measured review
effectiveness metrics (Section 9.4).

\hypertarget{acf-e1-implicit-privilege-grant}{%
\subsubsection{ACF-E1: Implicit Privilege
Grant}\label{acf-e1-implicit-privilege-grant}}

\textbf{STRIDE:} Elevation of Privilege \textbar{} \textbf{Risk:}
Critical \textbar{} \textbf{Detection:} None

\textbf{Description:} External system assertions are accepted without
independent verification, granting privileges based on unvalidated
claims.

\textbf{Why agents produce this:} Agents implement integration patterns
by calling external APIs and acting on the response. The concept that
the external system's response must be independently verified --- that
the response itself is untrusted --- is not visible in the code
structure. The code looks like a normal API call and response handling.

\textbf{Example:}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{partner\_verification }\OperatorTok{=}\NormalTok{ partner\_api.verify\_identity(applicant\_id)}
\ControlFlowTok{if}\NormalTok{ partner\_verification.get(}\StringTok{"verified"}\NormalTok{, }\VariableTok{False}\NormalTok{):}
\NormalTok{    grant\_system\_access(applicant\_id, level}\OperatorTok{=}\StringTok{"standard"}\NormalTok{)}
\CommentTok{\# Partner says "verified" → access granted.}
\CommentTok{\# No independent check. No recording of the basis for the decision.}
\CommentTok{\# If the partner system is compromised, every applicant is "verified."}
\end{Highlighting}
\end{Shaded}

\textbf{Detection approach:} Taint analysis --- the return value of an
\texttt{@external\_boundary} function is used as a predicate in an
access control decision without passing through validation. Requires
both boundary annotation and understanding of which operations are
access-control-relevant.

\textbf{Related:} ACF-S3 (structural identity spoofing) describes a
complementary elevation mechanism where the gate itself is structurally
unsound --- \texttt{hasattr()} gates accept any object that declares the
expected attribute, regardless of type. ACF-E1 is about
\emph{unvalidated external assertions}; ACF-S3 is about \emph{unsound
gate predicates}. Both result in implicit privilege grants, but through
different mechanisms.

\hypertarget{acf-e2-unvalidated-delegation}{%
\subsubsection{ACF-E2: Unvalidated
Delegation}\label{acf-e2-unvalidated-delegation}}

\textbf{STRIDE:} Elevation of Privilege \textbar{} \textbf{Risk:} High
\textbar{} \textbf{Detection:} Partial

\textbf{Description:} User-supplied parameters are used directly in
privileged operations (database queries, file access, system commands)
without validation or restriction.

\textbf{Why agents produce this:} The pattern
\texttt{db.query(Model).filter\_by(**user\_params)} is concise and
idiomatic. Agents produce it because it's the shortest path from input
to query. The concept that user parameters must be restricted to an
allowlist of permitted fields is a security requirement, not a language
requirement.

\textbf{Example:}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{def}\NormalTok{ search\_records(user\_query: }\BuiltInTok{dict}\NormalTok{):}
    \ControlFlowTok{return}\NormalTok{ db.query(Record).filter\_by(}\OperatorTok{**}\NormalTok{user\_query)}
    \CommentTok{\# User can filter on internal fields: is\_deleted, internal\_score,}
    \CommentTok{\# admin\_notes — fields that should not be queryable.}
\end{Highlighting}
\end{Shaded}

\textbf{Detection approach:} SQL injection scanners catch some cases
(especially string interpolation into SQL). Parameter delegation via
\texttt{**kwargs} unpacking into ORM queries is less consistently
detected. Semantic detection requires understanding which operations are
privileged and which parameters are user-controlled.

\hypertarget{detection-capability-summary}{%
\subsection{Detection Capability
Summary}\label{detection-capability-summary}}

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.3265}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.1429}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2653}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2653}}@{}}
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Detection Level
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Count
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Failure IDs
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Implication
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
\textbf{None} (no existing tool detects it) & 5 & ACF-S1, ACF-T1,
ACF-T2, ACF-R2, ACF-E1 & These require new tooling or new review
practices \\
\textbf{Partial} (some tools catch some cases) & 5 & ACF-S2, ACF-S3,
ACF-R1, ACF-I1, ACF-E2 & Existing tools provide incomplete coverage;
augmentation needed \\
\textbf{Good} (existing tools generally catch it) & 1 & ACF-I2 & Already
addressed by current tooling \\
\textbf{N/A} (process threat, not code pattern) & 2 & ACF-D1, ACF-D2 &
Requires process controls, not technical controls \\
\end{longtable}

This distribution --- 5 of 13 failure modes completely undetectable by
existing tools, and 5 more only partially detected --- is the gap this
paper identifies. The 5 undetectable failure modes include both
Critical-rated entries (ACF-T1, ACF-E1), meaning the highest-risk
failures are precisely the ones that current tooling misses entirely.

\hypertarget{appendix-b-technical-feasibility-of-automated-enforcement}{%
\section{Appendix B: Technical Feasibility of Automated
Enforcement}\label{appendix-b-technical-feasibility-of-automated-enforcement}}

\hypertarget{b.1-design-history}{%
\subsection{B.1 Design History}\label{b.1-design-history}}

The threat model in this paper identifies failure modes that existing
tools don't detect (Appendix A: 5 of 13 failure modes have no existing
tool coverage). The natural question is: can these be detected
automatically?

The design presented in this appendix emerged through three iterations
of structured adversarial deliberation using prompted AI agent teams,
spanning several days. The iterative history is documented here because
it is itself evidence for the paper's thesis: agent output treated as
external input (Section 5.2), validated through structured review
(Section 7.1), with the human providing framing and acceptance judgment
rather than technical content.

\textbf{Iteration 1: ``Can we fix Python?''} The initial framing asked
whether Python's permissive defaults could be addressed at the language
level --- a stricter dialect, a transpiler, or a runtime enforcement
layer. An agent team deliberated across five rounds and concluded
unanimously that the approach was not viable. The ecosystem orphaning
risk (a Python variant diverges from CPython with every release), the
adoption cliff, and the maintenance burden were each independently
fatal.

\textbf{Iteration 2: ``What can we do instead?''} The human contribution
at this stage was a question, not a direction. A second agent team ---
operating with knowledge of the first team's conclusions but no other
constraint --- independently generated the concept that became the
semantic boundary enforcer: a standalone AST-based analysis layer that
extends Python's existing annotation machinery rather than creating a
parallel system. This team conducted eight rounds of structured
adversarial deliberation and produced the initial design specification.
The design was implemented as a pattern-matching enforcement gate and
deployed in production on the case study codebase (Section 8), where it
catches the 1--2 violations per day reported in Section 8.4.

The creative pivot --- from ``change the language'' to ``analyse the
language'' --- was generated by the agent team, not by the human
operator. The human asked an open question; the agents produced the core
insight that the problems identified in Iteration 1 are analysable over
standard Python using its existing annotation system. The human
contribution was the question, the validation structure, and the
judgment to proceed --- not the design concept itself.

That existing tool operates by matching known defensive anti-patterns
(\texttt{.get()} with defaults on typed data, broad \texttt{except}
clauses on audit-critical paths, \texttt{hasattr()} usage) against a
per-module allowlist. It is effective for the patterns it recognises,
but it has structural limitations: it cannot trace data flow across
function boundaries, it cannot verify that a function decorated as a
validator actually contains control flow, and it cannot distinguish
contextually appropriate uses of a pattern from dangerous ones without
per-finding allowlist entries.

\textbf{Iteration 3: ``What does the best version look like?''} The
third iteration was designed as a reproducibility test --- a clean-start
roundtable with the same problem statement. It used seven specialist
agent perspectives (security architect, Python AST engineer, systems
thinker, adversarial red-teamer, quality engineer, governance designer,
integration engineer) with a dedicated scribe agent maintaining shared
working memory between rounds. The scribe had authority to challenge
agents who contradicted their prior stated positions --- functioning as
a committee chair enforcing epistemic accountability.

The process included steelman rounds (each agent argues for the
\emph{strongest} version of the position they most disagree with) and
structured challenge rounds (independent fault-finding). These
mechanisms produced measurably different results from single-pass
review: binary taint tracking was rejected 7/7 after cross-agent
challenge exposed it as a compliance ritual; runtime trust tagging was
rejected after four of seven agents independently converged on the same
fatal objection; and the team unanimously rejected the human-agent dual
enforcement model from Iteration 2 on grounds that authorship
attribution is unsolvable in mixed workflows --- an argument the
original roundtable never surfaced.

The reproducibility finding was nuanced: the process dynamics reproduced
(convergent rejections, forced concessions), but specific architectural
decisions diverged. Most significantly, the team replaced binary taint
with a two-dimensional model tracking provenance (where data came from)
and validation status (what processing it received) as orthogonal
dimensions --- an insight two agents independently converged on in
parallel, which constitutes stronger evidence for its validity than a
single agent proposing it. The design presented in this appendix
reflects the third iteration's architecture, noting where it extends the
production implementation.

\textbf{The human role across iterations.} In Iteration 1, the human
posed the initial question. In Iteration 2, the human posed the
follow-up question, validated the agent-generated concept, and judged
when the design was sufficient for implementation. In Iteration 3, the
human designed the reproducibility experiment and will ultimately judge
whether the more sophisticated design warrants implementation. At no
stage did the human generate the core technical insight --- but at every
stage, the human determined which agent outputs entered the trusted
store. This is the paper's proposed workflow (Section 5.3) operating at
the design level.

\textbf{Feasibility finding.} The existing production deployment
demonstrates that automated detection of the most critical agentic code
failure modes is technically feasible for Python, buildable at modest
cost, and compatible with existing development workflows --- even with
the relatively simple pattern-matching approach of Iteration 2. The
enhanced design presented here extends that foundation with
provenance-aware taint tracing and structural verification, addressing
the limitations above. It is presented so that other organisations can
build equivalent tooling for their own codebases and languages ---
whether the approach generalises beyond Python is an open question (see
Recommendation 4). The full roundtable synthesis --- including the
complete rule evaluation matrix, governance model, and minority reports
--- is available as a companion document.

\hypertarget{b.2-core-properties}{%
\subsection{B.2 Core Properties}\label{b.2-core-properties}}

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.5556}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.4444}}@{}}
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Property
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Detail
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
\textbf{Dependencies} & Zero --- Python standard library \texttt{ast}
module only \\
\textbf{Language compatibility} & Analyses standard Python; no custom
syntax or runtime modifications \\
\textbf{Output format} & SARIF (Static Analysis Results Interchange
Format) \\
\textbf{Delivery} & Standalone PyPI package \\
\textbf{Taint model} & Two-dimensional: provenance (where data came
from) × validation status (what processing it received) --- 7 effective
states determining finding severity \\
\textbf{Analysis scope} & Intra-function taint analysis (v0.1);
inter-procedural deferred to v1.0 \\
\end{longtable}

Note: the relevant measure of this tool is not its implementation size
but its \textbf{verification properties} --- see Section B.5.

\hypertarget{b.3-how-it-works}{%
\subsection{B.3 How It Works}\label{b.3-how-it-works}}

\textbf{Declaration.} Developers declare trust boundaries using standard
Python annotations and decorators --- no custom syntax, no new imports
beyond the tool's own library:

\begin{Shaded}
\begin{Highlighting}[]
\AttributeTok{@external\_boundary}          \CommentTok{\# Returns untrusted data}
\KeywordTok{def}\NormalTok{ fetch\_api\_data(): ...}

\AttributeTok{@validates\_external}         \CommentTok{\# Validates external data (must contain control flow)}
\KeywordTok{def}\NormalTok{ validate\_response(): ...}
\end{Highlighting}
\end{Shaded}

A built-in heuristic list recognises common external call sites
(\texttt{requests.*}, \texttt{httpx.*}, \texttt{sqlalchemy.*.execute},
\texttt{json.loads}, etc.) without requiring explicit annotation ---
reducing the annotation burden for common patterns while allowing
project-specific declarations for uncommon ones.

\textbf{Two-dimensional taint model.} Unlike binary taint tracking
(tainted/clean), the tool tracks two independent dimensions for every
variable in scope. \emph{Provenance} --- where the data came from --- is
immutable once assigned: \texttt{TIER\_1} (audit trail),
\texttt{TIER\_2} (pipeline data), \texttt{TIER\_3} (external),
\texttt{UNKNOWN}, or \texttt{MIXED} (container holding values from
multiple tiers). \emph{Validation status} --- what processing the data
received --- monotonically increases from \texttt{RAW} to
\texttt{STRUCTURALLY\_VALIDATED}. Validation is only meaningful for
\texttt{TIER\_3} and \texttt{UNKNOWN} data; internal data does not pass
through \texttt{@validates\_external}. This reduces the theoretical 5×2
matrix to 7 effective states, each producing different finding severity
for the same pattern. For example, \texttt{.get()} with a default on
\texttt{TIER\_1} data is an unconditional error (audit trail
fabrication), on \texttt{TIER\_2} data is a standard error (contract
violation), and on validated \texttt{TIER\_3} data is suppressed (the
validation boundary already handled it).

\textbf{Analysis.} The tool performs two-pass analysis:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  \textbf{Symbol collection:} Walks the AST to build a table of
  annotated boundaries, decorated functions, and known external call
  sites
\item
  \textbf{Rule evaluation:} Seven rules detect defensive programming
  patterns that are dangerous in high-assurance contexts ---
  \texttt{.get()} with defaults, \texttt{getattr()} with fallbacks,
  \texttt{hasattr()} gates, broad \texttt{except} blocks, silent
  exception swallowing, unvalidated data reaching audit writes, and
  \texttt{isinstance()} guards on internal data. Each rule is evaluated
  against the 7 effective states, producing a 49-cell severity matrix
  where every combination of pattern and provenance has a defined
  outcome (error, warning, advisory, or suppress). The full matrix is
  documented in the companion roundtable synthesis.
\end{enumerate}

\textbf{Structural verification.} Functions decorated with
\texttt{@validates\_external} must contain a rejection path in their
body --- reachable control flow that can reject invalid data
(\texttt{try/except}, \texttt{isinstance()} with conditional,
\texttt{raise}, or \texttt{if/else}). A validator without a rejection
path is structurally unsound --- it attests to validation without the
ability to reject. This prevents agents from satisfying the taint
checker by applying a decorator to a pass-through function.

\hypertarget{b.4-governance-model}{%
\subsection{B.4 Governance Model}\label{b.4-governance-model}}

The tool addresses the agent circumvention problem (Section 7 of the
threat model) through three mechanisms, all of which are implemented and
enforced in the production deployment described in Section 8:

\textbf{Project-level manifest.} A \texttt{strict.toml} file declares
trust topology, boundary functions, rule configuration, and structured
exceptions. Exceptions require a rationale, reviewer identity, and
expiry date.

\textbf{Human-gated governance.} The manifest is protected by repository
CODEOWNERS. Ratification of new exceptions must occur in a
\textbf{separate, prior commit} from the source code changes --- CI
verifies the temporal separation via git commit history. This prevents
agents from simultaneously generating a trust boundary violation and its
exception in the same commit.

\textbf{Exceptionability classification.} Each cell in the 49-cell
severity matrix is assigned to one of four governance classes:
\emph{UNCONDITIONAL} (no exceptions permitted --- the tool rejects
creation at parse time), \emph{STANDARD} (grouped rationale with 90-day
expiry and divergence detection), \emph{LIBERAL} (single-line rationale
with 180-day expiry), or \emph{TRANSPARENT} (advisory or suppressed ---
below governance threshold). Twenty-four of 49 cells (49\%) are
UNCONDITIONAL --- project invariants hardcoded in the tool's source, not
in editable configuration. These include all \texttt{hasattr()} findings
regardless of provenance, \texttt{.get()} on audit trail data, and broad
exception handlers on audit-critical paths. An earlier iteration of the
design used differential enforcement profiles for human-authored and
agent-authored code; this was unanimously rejected on the grounds that
authorship attribution is unsolvable in mixed workflows and that
differential profiles create incentives to launder agent code through
human commits.

\hypertarget{b.5-verification-properties-the-assurance-argument}{%
\subsection{B.5 Verification Properties (The Assurance
Argument)}\label{b.5-verification-properties-the-assurance-argument}}

For a policy audience, the relevant question about this tool is not
``how big is it?'' but \textbf{``how do you know it's correct?''} This
is especially pertinent if the tool is itself built with agent
assistance --- the tool's own development is subject to the threat model
it exists to address.

The design includes four independent verification mechanisms:

\textbf{1. Golden corpus.} A collection of labelled Python snippets ---
true positives (code that should trigger findings) and true negatives
(code that should not) --- plus adversarial evasion samples (code that
looks compliant but isn't). Minimum: 3 true positives + 2 true negatives
per rule. The corpus is a first-class artefact, version-controlled, and
a ship gate.

\textbf{2. Self-hosting gate.} The tool's own source code must pass its
own rules in CI from the first commit. If the tool cannot be written to
its own standards, the standards are not understood.

\textbf{3. Measured precision with volume floor.} Rules track their true
positive rate across runs. An immutable 80\% precision floor is
hardcoded in the tool --- below this, a rule cannot earn blocking status
regardless of configuration. Per-rule thresholds are earned above this
floor through corpus evidence (e.g., \texttt{hasattr()} at
\textasciitilde99\%, \texttt{.get()} at \textasciitilde88\%) and are
monotonically non-decreasing: once a rule earns a higher threshold, it
cannot be lowered. This prevents both promotion based on small samples
and regression of established precision.

\textbf{4. Deterministic output.} Identical input must produce
byte-identical output. Verified by running the tool twice and diffing.
Non-determinism in a security tool is a defect.

These properties are \textbf{independently evaluable by an assessor.} An
IRAP assessor (or equivalent) can verify:

\begin{itemize}
\tightlist
\item
  The corpus exists and covers the claimed rules
\item
  The self-hosting gate is enforced in CI
\item
  Precision tracking data is available and shows the claimed rates
\item
  Output is deterministic
\end{itemize}

This is a stronger assurance argument than ``the implementation is N
lines'' or ``the development took M hours.'' It answers the question
\emph{``how do you know the tool catches what it claims to catch?''}
with measurable, auditable evidence.

\hypertarget{b.6-relationship-to-existing-tools}{%
\subsection{B.6 Relationship to Existing
Tools}\label{b.6-relationship-to-existing-tools}}

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2500}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2321}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2857}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2321}}@{}}
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Existing Tool
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
What It Does
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
What It Misses
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Relationship
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
\textbf{mypy/pyright} & Verifies type shape & Data provenance (a
\texttt{str} from an API and a \texttt{str} from a DB are the same type)
& Complementary --- this tool verifies provenance \\
\textbf{ruff/flake8} & Enforces style/idiom & Trust boundary semantics &
Complementary --- this tool handles trust boundaries; style belongs in
ruff \\
\textbf{bandit} & Detects known vulnerability patterns &
Project-specific trust violations & Complementary --- bandit is generic;
this tool is project-configured \\
\textbf{semgrep} & Custom pattern matching with taint analysis &
Provenance-aware severity grading, exceptionability classification,
manifest-based exception governance & Closest existing tool --- but
lacks the governance model \\
\end{longtable}

\textbf{Why not extend an existing tool?} The governance model
(exceptionability classification with UNCONDITIONAL cells hardcoded in
source, provenance-aware severity grading, temporal separation of
manifest changes) is the differentiating capability. These are not
features that can be added to existing tools as plugins --- they require
control over the finding lifecycle, which existing tools' architectures
don't expose. The analysis engine is deliberately simple; the governance
model is the novel contribution.

\hypertarget{b.7-the-meta-observation}{%
\subsection{B.7 The Meta-Observation}\label{b.7-the-meta-observation}}

If this tool is built with agent assistance --- which is the natural
development approach given the domain --- its own development becomes
the first test case for its thesis. If the design specification is tight
enough that an agent can implement it, and the self-hosting gate and
golden corpus catch the problems the agent introduces, the tool has
demonstrated its value proposition through its own creation.

This recursive property is unusual for security tools and is worth
noting for the policy audience: \textbf{the tool's development process
is itself evidence for or against its claims.} An assessor can evaluate
not just the tool's output but the conditions under which it was built
--- did the self-hosting gate catch agent-generated defects? Did the
golden corpus reveal false positives or negatives? The development
history is part of the assurance argument.

The full design specification is available as a companion document:
\emph{Semantic Boundary Enforcer --- Design Specification}
(forthcoming).

\hypertarget{appendix-c-agent-autonomy-self-assessment}{%
\section{Appendix C: Agent Autonomy
Self-Assessment}\label{appendix-c-agent-autonomy-self-assessment}}

This appendix provides an informal diagnostic for organisations to
identify where their current agent usage sits on the autonomy spectrum
and whether their controls are proportionate. It is not a maturity model
--- higher tiers are not aspirational targets, and there is no implied
progression from lower to higher. Most organisations will find
themselves at different tiers simultaneously: Tier 1 for
security-critical components, Tier 3 for test scaffolding, Tier 2 for
general feature work. That is entirely appropriate, provided the
controls at each tier match the risk profile described below.

The purpose is self-location. An organisation that discovers it is
operating at Tier 3 without the controls listed for Tier 3 has
identified a gap. An organisation operating at Tier 1 with controls
designed for Tier 3 has identified waste. Neither outcome requires
changing the tier --- only aligning controls to reality.

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2000}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2000}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2000}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2000}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2000}}@{}}
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Tier 0: Full Human}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Tier 1: Prompted + Copied}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Tier 2: IDE-Integrated}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Tier 3: Autonomous}
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
\textbf{What it looks like} & No agent involvement. Human writes all
code. & Developer asks agent specific questions, copies and adapts
fragments into codebase manually. & Agent autocompletes functions and
classes inline. Developer accepts or rejects suggestions in-editor. &
Agent plans, implements, tests, and commits with minimal human
intervention. \\
\textbf{Who holds architectural context} & Human & Human & Shared ---
human directs, agent infers from surrounding code & Agent --- from
project documentation, system prompts, and codebase patterns \\
\textbf{Error correlation} & Independent (human variation) & Low ---
fragments are isolated, human integrates & Moderate --- agent infers
patterns from local context and may replicate across completions & High
--- same training biases applied systematically across features (Section
2.4) \\
\textbf{Review surface} & Normal & Slightly elevated --- more code to
review, but each fragment is small & Elevated --- easy to accept
completions without full evaluation & Massive --- entire features arrive
at review boundary as finished artefacts (Section 4.1) \\
\textbf{Habituation risk} & Baseline & Low & Moderate --- ``tab-accept''
becomes reflexive (Section 4.2) & High --- output volume degrades review
from verification to scanning \\
\textbf{Minimum controls} & Existing SDLC & Existing SDLC is likely
adequate & Awareness of ACF patterns (Appendix A); SAST augmentation
advisable & Validation boundary (Section 5.3), semantic enforcement
(Section 7.2), provenance tracking (Section 7.3), measured review
effectiveness (Section 9.4) \\
\end{longtable}

\hypertarget{references}{%
\section{References}\label{references}}

\begin{itemize}
\tightlist
\item
  Australian Signals Directorate. \emph{Information Security Manual.}
  Commonwealth of Australia. December 2025 revision.
  \url{https://www.cyber.gov.au/resources-business-and-government/essential-cyber-security/ism}
  --- Controls referenced: ISM-0401 (Secure by Design), ISM-0402
  (SAST/DAST/SCA), ISM-1419 (development environments),
  ISM-2026/2027/2028 (software artefact integrity), ISM-2060 (code
  review), ISM-2061 (security-focused peer review), ISM-2074 (AI usage
  policy). Individual controls are searchable by number on the ASD
  website.
\item
  Australian Signals Directorate. \emph{Essential Eight Maturity Model.}
  Commonwealth of Australia. Updated periodically.
  \url{https://www.cyber.gov.au/resources-business-and-government/essential-cyber-security/essential-eight}
\item
  NIST. \emph{SP 800-218: Secure Software Development Framework (SSDF),
  Version 1.1.} February 2022.
  \url{https://csrc.nist.gov/publications/detail/sp/800-218/final}
\item
  NIST. \emph{SP 800-218A: Secure Software Development Practices for
  Generative AI and Dual-Use Foundation Models.} July 2024.
  \url{https://csrc.nist.gov/publications/detail/sp/800-218a/final} ---
  (AI-specific SSDF supplement referenced in Section 6.2)
\item
  Microsoft. \emph{The STRIDE Threat Model.} Microsoft Security
  Development Lifecycle.
  \url{https://learn.microsoft.com/en-us/azure/security/develop/threat-modeling-tool-threats}
\item
  OASIS. \emph{Static Analysis Results Interchange Format (SARIF),
  Version 2.1.0.} March 2020.
  \url{https://docs.oasis-open.org/sarif/sarif/v2.1.0/sarif-v2.1.0.html}
  --- (Output format referenced in Appendix B)
\item
  OWASP. \emph{Top 10 for LLM Applications.} 2025.
  \url{https://owasp.org/www-project-top-10-for-large-language-model-applications/}
\item
  OWASP. \emph{Secure Coding Practices --- Quick Reference Guide.} 2010.
  \url{https://owasp.org/www-project-secure-coding-practices-quick-reference-guide/}
\item
  MITRE. \emph{Common Weakness Enumeration (CWE).} The MITRE
  Corporation. \url{https://cwe.mitre.org/} --- (Taxonomy comparison in
  Section 6.4)
\item
  Meadows, D. \emph{Thinking in Systems: A Primer.} Chelsea Green
  Publishing, 2008. (Systems archetypes referenced in Section 4.2)
\item
  Parasuraman, R. \& Manzey, D. ``Complacency and Bias in Human Use of
  Automation: An Attentional Integration.'' \emph{Human Factors}, 52(3),
  381-410, 2010. (Automation bias referenced in Section 4.2)
\item
  Perry, N. et al.~``Do Users Write More Insecure Code with AI
  Assistants?'' \emph{ACM CCS}, 2023. (AI-assisted developers wrote less
  secure code while feeling more confident --- Section 4.2)
\item
  Peng, S. et al.~``The Impact of AI on Developer Productivity: Evidence
  from GitHub Copilot.'' 2023. (Controlled study: 55.8\% faster task
  completion --- Section 1.2.1)
\item
  Cui, K. et al.~``The Effects of Generative AI on High Skilled Work:
  Evidence from Three Field Experiments with Software Developers.''
  MIT/Microsoft/Accenture, 2024. (Field experiment: 12.9--21.8\% more
  PRs/week at Microsoft --- Section 1.2.1)
\item
  Dutta, P. et al.~``ASTRIDE: A Security Threat Modeling Platform for
  Agentic-AI Applications.'' \emph{arXiv:2512.04785}, December 2025.
  \url{https://arxiv.org/abs/2512.04785} --- (STRIDE extension for AI
  agent-specific attacks, referenced in Section 3.1)
\item
  Meier, J.D. et al.~``STRIDE-LM Threat Model.'' \emph{CSF Tools}.
  \url{https://csf.tools/reference/stride-lm/} --- (STRIDE extension
  adding Lateral Movement, referenced in Section 3.1)
\item
  METR. ``Measuring the Impact of Early-2025 AI on Experienced
  Open-Source Developer Productivity.'' July 2025. (RCT: experienced
  developers 19\% slower with AI despite believing 20\% faster ---
  Sections 1.2.1, 4.2)
\item
  Pichai, S. Alphabet Inc.~Q3 2024 Earnings Call, 29 October 2024.
  ``More than a quarter of all new code at Google is generated by AI,
  then reviewed and accepted by engineers.'' Reported in \emph{The
  Verge}, 29 October 2024.
  \url{https://www.theverge.com/2024/10/29/24282757/google-new-code-generated-ai-q3-2024}
  --- (Section 1.2.1)
\item
  Wolf, A. ``Welcome to the Eternal September of open source. Here's
  what we plan to do for maintainers.'' \emph{GitHub Blog}, 12 February
  2026 (updated 13 February 2026).
  \url{https://github.blog/open-source/maintainers/welcome-to-the-eternal-september-of-open-source-heres-what-we-plan-to-do-for-maintainers/}
  --- Primary source. GitHub's Director of Open Source Programs on the
  review capacity crisis: ``The cost to create has dropped but the cost
  to review has not.'' Announces PR access controls, interaction limits,
  and automated triage. (Referenced in Sections 1.2.2 and 4.1)
\item
  Ghoshal, A. ``GitHub eyes restrictions on pull requests to rein in
  AI-based code deluge on maintainers.'' \emph{InfoWorld}, 4 February
  2026.
  \url{https://www.infoworld.com/article/4127156/github-eyes-restrictions-on-pull-requests-to-rein-in-ai-based-code-deluge-on-maintainers.html}
  --- Secondary reporting on the same events. GitHub described the
  problem as a denial-of-service attack on human attention. (Review
  capacity exhaustion evidence referenced in Sections 1.2.2 and 4.1)
\item
  Graham-Cumming, J. ``Incident report on memory leak caused by
  Cloudflare parser bug.'' Cloudflare Blog, 23 February 2017.
  \url{https://blog.cloudflare.com/incident-report-on-memory-leak-caused-by-cloudflare-parser-bug/}
  --- (Precedent referenced in Section 1.2)
\end{itemize}

\textbf{Suggested citation:} Morrissey, J. (ORCID:
\href{https://orcid.org/0009-0000-5654-3782}{0009-0000-5654-3782}).
``When Agents Write Code: A Threat Model for AI-Assisted Software
Development in Government Systems.'' Discussion Paper, DRAFT v0.2, 8
March 2026. Digital Transformation Agency.

\end{document}
